---
title: "Getting started with cppally"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cppally}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

```{r, include = FALSE}
required <- c("bench", "brio", "callr", "cli", "decor",
              "desc", "glue", "purrr", "readr", "stringr",
              "utils", "vctrs", "withr")
if (!all(vapply(required, requireNamespace, logical(1), quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
  knitr::knit_exit()
}
```

Below we explore some of the capabilities of cppally, from its custom C++ scalar and vectors, to using templates and concepts.

**Note:** The classes `r_vec` and `r_vector` are aliases of one another and thus can be used interchangeably.

## Setup

Let's start by loading cppally

```{r}
library(cppally)
```

```{r, include=FALSE}
# Helpers to compile all examples in debug mode
cpp_source <- function(..., code, debug = TRUE, env = parent.frame()){
  preamble <- c("#include <cppally.hpp>", "using namespace cppally;")
  code <- paste(c(preamble, code), collapse = "\n")
  cppally::cpp_source(debug = debug, env = env, code = code, ...)
}
cpp_eval <- function(..., debug = TRUE, env = parent.frame()){
    cppally::cpp_eval(debug = debug, env = env, ...)
}

# Helpers to source and display C++/R code
chunk_impl <- function(x, language){
  paste0("```", language, "\n", x, "\n```\n")
}
as_code_chunk <- function(x, language){
  cat(chunk_impl(x, language))
}
as_cpp_chunk <- function(x){
  as_code_chunk(x, "cpp")
}

# Pre-register named single-line expressions so they can be referenced later
register_single_exprs <- function(exprs, env = parent.frame(), ...){
  if (is.null(names(exprs))){
    stop("`exprs` must be named")
  }
  utils::getFromNamespace("source_single_exprs", "cppally")(
    exprs, env = env, ...
    )
  wrappers <- setNames(
    lapply(seq_along(exprs), \(i) {
      fn <- get(paste0("f", i), envir = env)
      function() {
        out <- fn()
        if (out[["is_void"]]) invisible() else out$result
      }
    }),
    names(exprs)
  )
  list2env(wrappers, envir = env)
  invisible()
}
```

```{r, include=FALSE}
# Compile necessary examples in one-go
# as it's faster when building the vignette

examples <- c(
  hello_world = '
[[cppally::register]]
void hello_world(){
  print("Hello World!");
}',
  lgl_ops = '
[[cppally::register]]
r_vector<r_lgl> lgl_ops(){
  return make_vec<r_lgl>(
    r_true || r_false, // true
    r_true && r_false, // false
    r_na || r_true,    // true
    r_na && r_true,    // NA
    r_na && r_false,   // false
    r_na || r_na,      // NA
    r_na && r_na      // NA
  );
}
',
  bad_lgl_print = '
[[cppally::register]]
void bad_lgl_print(r_lgl condition){
  if (condition){
    print("true");
  } else {
    print("false");
  }
}
',
  good_lgl_print = '
[[cppally::register]]
void good_lgl_print(r_lgl condition){
  if (is_na(condition)){
    print("NA");
  } else if (condition){
    print("true");
  } else {
    print("false");
  }
}
',
  also_good_lgl_print = '
[[cppally::register]]
void also_good_lgl_print(r_lgl condition){
  if (condition.is_true()){
    print("true");
  } else {
    print("not true");
  }
}
',
cppally_equality = '
[[cppally::register]]
void cppally_equality(){

  r_int x = na<r_int>();
  r_int y = na<r_int>();
  
  r_lgl x_equal_to_y = x == y;
  bool x_identical_to_y = identical(x, y);
  
  // NA so not printed
  if ( x_equal_to_y.is_true() ){
    print("x is equal to y\\n");
  }
  
  // NA so not printed
  if ( x_equal_to_y.is_false() ){
    print("x is not equal to y\\n");
  }
  
  // NA so printed
  if (is_na(x_equal_to_y)){
    print("`x == y` produces `NA`\\n");
  }
  
  // Both na<r_int>() therefore they are identical to each other
  if (x_identical_to_y){
    print("x is identical to y\\n");
  }
}
',
cppally_identical = '
template <typename T, typename U>
[[cppally::register]]
bool cpp_identical(T x, U y){
  return identical(x, y);
}
',
  new_integer_vector = '
// Integer vector of size n
[[cppally::register]]
r_vector<r_int> new_integer_vector(int n){
  r_vector<r_int> int_vctr(n, /*fill = */ r_int(0));
  return int_vctr;
}
',
  all_vectors = '
[[cppally::register]]
r_vector<r_sexp> all_vectors(){
  return make_vec<r_sexp>(
    arg("logical") = r_vector<r_lgl>(),
    arg("integer") = r_vector<r_int>(),
    arg("integer64") = r_vector<r_int64>(), // Requires bit64
    arg("double") = r_vector<r_dbl>(),
    arg("character") = r_vector<r_str>(),
    arg("character") = r_vector<r_str_view>(),
    arg("raw") = r_vector<r_raw>(),
    arg("date") = r_vector<r_date>(),
    arg("date-time") = r_vector<r_psxct>(),
    arg("list") = r_vector<r_sexp>()
  );
}
',
cppally_math = '
[[cppally::register]]
r_vector<r_dbl> cppally_math(r_dbl x){
    return make_vec<r_dbl>(
        arg("abs")          = abs(x),
        arg("floor")        = floor(x),
        arg("ceiling")      = ceiling(x),
        arg("trunc")        = trunc(x),
        arg("round")        = round(x),
        arg("signif")       = signif(x, 3),
        arg("sign")         = sign(x),
        arg("min")          = min(0, x),
        arg("max")          = max(0, x),
        arg("sqrt")         = sqrt(x),
        arg("pow")          = pow(x, 2),
        arg("exp")          = exp(x),
        arg("log")          = log(x),
        arg("log_base")     = log(x, 2),
        arg("log10")        = log10(x)
    );
}', 
  cpp_abs = '
template <RNumber T>
[[cppally::register]]
T cpp_abs(T x){
  if (is_na(x)){
    return na<T>();
  } else if (x < 0){
    return -x;
  } else {
    return x;
  }
}
',
  scalar_default = '
// Return the default constructor result of RScalar types

template <RScalar T>
[[cppally::register]]
T scalar_default(T ptype){
	return T();
}
',
  double_to_int = '
[[cppally::register]]
r_int double_to_int(r_dbl x){
  return as<r_int>(x);
}
',
  to_int_vec = '
[[cppally::register]]
r_vector<r_int> to_int_vec(r_vector<r_dbl> x){
  return as<r_vector<r_int>>(x);
}
',
  coercions = '
[[cppally::register]]
r_vector<r_sexp> coercions(){
    r_dbl a(4.2);
    r_vector<r_dbl> b = make_vec<r_dbl>(2.5);
    return make_vec<r_sexp>(
        as<r_vector<r_int>>(a),
        as<r_int>(a),
        as<r_int>(b),
        as<r_dbl>(b)
    );
}
',
to_from_cpp_vec = '
[[cppally::register]]
r_vector<r_int> cpp_vectors_example(r_vector<r_int> x){
  std::vector x_cpp = as<std::vector<r_int>>(x);
  x_cpp.push_back(r_int(42));
  return as<r_vector<r_int>>(x_cpp);
}
',
  str_concatenate = '
[[cppally::register]]
r_str str_concatenate(r_str x, r_str y, r_str sep){
  std::string left = std::string(x.cpp_str());
  std::string right = std::string(y.cpp_str());
  std::string middle = std::string(sep.cpp_str());
  std::string combined = left + middle + right;
  return r_str(combined.c_str());
}
',
  new_list = '
using list = r_vector<r_sexp>;

[[cppally::register]]
list new_list(int n){
  return list(n);
}
',
  resize_all = '
[[cppally::register]]
r_vector<r_sexp> resize_all(r_vector<r_sexp> x, r_size_t n){
    r_size_t list_length = x.length();
    for (r_size_t i = 0; i < list_length; ++i){
        r_sexp_visit(x.view(i), [&]<RVector T>(T vec) {
            x.set(i, vec.resize(n));
        });
    }
    return x;
}
',
  new_factor = '
[[cppally::register]]
r_factors new_factor(r_vector<r_str> x){
	return r_factors(x);
}
',
  factor_codes = '
static_assert(!RVector<r_factors>);

[[cppally::register]]
r_vector<r_int> factor_codes(r_factors x){
	return x.codes();
}
',
find_empty_string = '
[[cppally::register]]
r_vector<r_int> find_empty_string(r_vector<r_str> x){
    return x.find(r_str(""));
}
',
match_strs = '
[[cppally::register]]
r_vector<r_int> match_strs(r_vector<r_str> x, r_vector<r_str> table){
    return match(x, table);
}
',
cpp_in = '
[[cppally::register]]
r_vector<r_lgl> cpp_in(r_vector<r_str> x, r_vector<r_str> table){
    return x IS_IN table;
}
',
cpp_not_in = '
[[cppally::register]]
r_vector<r_lgl> cpp_not_in(r_vector<r_str> x, r_vector<r_str> table){
    return !(x IS_IN table);
}
',
cpp_subset = '
template <RVector T, typename subscript_t>
requires any<subscript_t, r_lgl, r_int, r_str>
[[cppally::register]]
T cpp_subset(T x, r_vector<subscript_t> y){
    return subset(x, y);
}
',
cpp_negative_subset = '
template <RVector T, typename subscript_t>
requires any<subscript_t, r_lgl, r_int, r_str>
[[cppally::register]]
T cpp_negative_subset(T x, r_vector<subscript_t> y){
    return subset(x, y, /*invert=*/ true);
}
',
  setnames = '
[[cppally::register]]
r_vector<r_sexp> set_list_names(r_vector<r_sexp> x, r_vector<r_str> names){
  x.set_names(names);
  return x;
}
',
cpp_seq_len = '
[[cppally::register]]
r_vector<r_int> cpp_seq_len(r_size_t n){
  return sequence(n, /* from = */ r_int(1), /* by = */ r_int(1));
}
',
cpp_sequences = '
template <typename T>
requires (any<T, r_int, r_int64, r_dbl>)
[[cppally::register]]
r_vector<r_sexp> cpp_sequences(r_vector<r_int> size, r_vector<T> from, r_vector<T> by){
    return pmap([](auto a, auto b, auto c){
        return as<r_sexp>(sequence(a, b, c));
    }, size, from, by);
}
',
get_time = '
[[cppally::register]]
r_date get_today(){
  return r_date::today();
}
[[cppally::register]]
r_psxct get_now(){
  return r_psxct::now();
}
',

get_time_components = '
[[cppally::register]]
r_int get_year(r_date x){
  return x.year();
}
[[cppally::register]]
r_int get_month(r_date x){
  return x.month();
}
[[cppally::register]]
r_int get_week(r_date x){
  return x.week();
}
[[cppally::register]]
r_int get_day(r_date x){
  return x.day();
}
[[cppally::register]]
r_int get_wday(r_date x, int week_start){
  return x.wday(week_start);
}
[[cppally::register]]
r_int get_yday(r_date x){
  return x.yday();
}
[[cppally::register]]
r_int get_iso_week(r_date x){
  return x.iso_week();
}
[[cppally::register]]
r_int get_iso_year(r_date x){
  return x.iso_year();
}
',
time_round = '
[[cppally::register]]
r_date floor_month(r_date x){
  return x.floor<"months">();
}
[[cppally::register]]
r_date ceiling_year(r_date x){
  return x.ceiling<"years">();
}
[[cppally::register]]
r_date round_week(r_date x, int week_start){
  return x.round<"weeks">(week_start);
}
',
create_date = '
[[cppally::register]]
r_date create_date(int year, int month, int day){
  return r_date(year, month, day);
}
[[cppally::register]]
r_date date_from_epoch_offset(double n){
  return r_date(n);
}

[[cppally::register]]
r_psxct create_datetime(int year, int month, int day, int hour, int minute, double second){
  return r_psxct(year, month, day, hour, minute, second);
}
[[cppally::register]]
r_psxct datetime_from_epoch_offset(double n){
  return r_psxct(n);
}
',
add_time = '
[[cppally::register]]
r_date add_days(r_date x, double n){
  return x.add<"days">(n);
}
[[cppally::register]]
r_date add_months(r_date x, double n){
  return x.add<"months">(n);
}
',
datetime_months = '
[[cppally::register]]
r_psxct as_dt(r_date x){
  return x.as_datetime();
}
[[cppally::register]]
r_psxct dt_add_months(r_psxct x, double n){
  return x.add<"months">(n);
}
[[cppally::register]]
r_dbl dt_diff_months(r_psxct x, r_psxct y){
  return time_diff<"months">(x, y);
}
',
time_diff = '
[[cppally::register]]
r_dbl diff_days(r_date x, r_date y){
  return time_diff<"days">(x, y);
}
[[cppally::register]]
r_dbl diff_months(r_date x, r_date y){
  return time_diff<"months">(x, y);
}
',
r_get_today = '
[[cppally::register]]
r_date r_get_today(){
  r_function sys_date("Sys.Date");
  return as<r_date>(sys_date());
}
',
base_sum_fn = '
[[cppally::register]]
r_function base_sum_fn(){
  return r_function(cached_sym<"sum">(), pkg_env<"base">());
}
',

base_sum = '
[[cppally::register]]
r_dbl base_sum(r_vector<r_dbl> x, bool na_rm){
  return as<r_dbl>(base_sum_fn()(x, arg("na.rm") = na_rm));
}
'
)

# Benchmarks need debug = FALSE
benchmark_examples <- c(
  cpp_unique = '
template <RVector T>
[[cppally::register]]
r_int cpp_n_unique(T x){
  return as<r_int>(n_unique(x));
}

template <RVector T>
[[cppally::register]]
T cpp_unique(T x, bool sort){
  return unique(x, sort);
}

',
cpp_sort = '

template <RSortableVector T>
[[cppally::register]]
r_vector<r_int> cpp_order(T x){
  
  // Add 1 to return 1-indexed permutation to match base::order()
  // In general C++ code, always use 0-indexed as cppally indexing is 
  // always 0-indexed.
  
  return order(x) + 1; 
}

template <RSortableVector T>
[[cppally::register]]
T cpp_sort(T x){
  return sort(x);
}
',

make_groups = '

template <RVector T>
[[cppally::register]]
r_vector<r_sexp> group_info(T x){
  groups g = make_groups(x, /*ordered=*/ false);
  return make_vec<r_sexp>(
    arg("group_id") = g.ids,
    arg("n_groups") = g.n_groups,
    arg("ordered") = g.ordered,
    arg("sorted") = g.sorted,
    arg("starts") = g.starts(),
    arg("counts") = g.counts(),
    arg("order") = g.order(),
    arg("group_names") = group_names(x, g)
  );
}
',
group_counts = '
template <RVector T>
[[cppally::register]]
r_df group_counts(T x){
  groups g = make_groups(x, /*ordered=*/ false);
  
  return make_df(
    arg("key") = subset(x, g.starts()),
    arg("count") = g.counts()
  );
}
',
cpp_stats = '
template <RNumber T>
[[cppally::register]]
r_dbl cpp_sum(r_vector<T> x, bool na_rm){
  return as<r_dbl>(sum(x, na_rm));
}
template <RNumber T>
[[cppally::register]]
r_vector<T> cpp_range(r_vector<T> x, bool na_rm){
  return range(x, na_rm);
}
template <RNumber T>
[[cppally::register]]
r_dbl cpp_var(r_vector<T> x, bool na_rm){
  return var(x, na_rm);
}
',
cpp_set_threads = '
[[cppally::register]]
void cpp_set_threads(int n){
  set_threads(n);
}

[[cppally::register]]
int cpp_get_threads(){
  return get_threads();
}
'
)

cpp_source(code = paste(examples, collapse = "\n"), debug = TRUE)
cpp_source(code = paste(benchmark_examples, collapse = "\n"), debug = FALSE)

# Single-line expressions, pre-registered as R functions of the same name.
# Each can be invoked later as e.g. `r_true_val()` to get the evaluated result.
single_exprs <- c(
  r_true_val          = 'r_true',
  r_false_val         = 'r_false',
  r_na_val            = 'r_na',
  r_na_check          = 'is_na(na<r_int>())',
  r_plus              = 'r_int(0) + r_dbl(2.5)',
  r_plus_na           = 'na<r_int>() + r_dbl(2.5)',
  r_minus             = 'r_int(0) - r_int(1)',
  r_minus_na          = 'na<r_int>() - r_int(1)',
  r_mult              = 'r_int(2) * r_int(3)',
  r_mult_na           = 'na<r_int>() * r_int(3)',
  r_div               = 'r_dbl(9) / 3',
  r_div_na            = 'na<r_dbl>() / 3',
  r_lt                = 'r_int(1) < r_int(2)',
  r_lt_na             = 'na<r_int>() < r_int(2)',
  r_lte                = 'r_dbl(2) <= r_dbl(2)',
  r_lte_na             = 'na<r_dbl>() <= r_dbl(2)',
  r_gt                = 'r_int(3) > r_int(2)',
  r_gt_na             = 'na<r_int>() > r_int(2)',
  r_gte                = 'r_dbl(2) >= r_dbl(3)',
  r_gte_na             = 'na<r_dbl>() >= r_dbl(3)',
  make_vec_dbl        = 'make_vec<r_dbl>(1, 1.5, 2, na<r_dbl>())',
  make_vec_dbl_named  = '
make_vec<r_dbl>(
    arg("first") = 1,
    arg("second") = 1.5,
    arg("third") = 2,
    arg("last") = na<r_dbl>()
  )
',
  make_vec_sexp       = 'make_vec<r_sexp>(1, 2, 3)',
  r_str_hello         = 'r_str("hello")',
  r_str_hello_c_str   = 'r_str("hello").c_str()',
  r_sym_new           = 'r_sym("new_symbol")',
  r_sym_from_str      = 'r_sym(r_str("symbol_from_string"))',
  cached_str_demo     = 'cached_str<"cached_string">()',
  cached_sym_demo     = 'cached_sym<"cached_symbol">()',
  cpp_seq             = 'seq(r_dbl(1), r_dbl(5), r_dbl(0.5))',
  cpp_sequence        = 'sequence(5, /*from = */ r_int(0), /*by = */ r_int(-1))'
)

register_single_exprs(single_exprs, debug = TRUE)
```

## Registering C++ functions (to R)

To make a C++ function available to R we use the `[[cppally::register]]` tag.

``` cpp
#include <cppally.hpp>
using namespace cppally;

[[cppally::register]]
void hello_world(){
  print("Hello World!");
}
```

After tagging our functions we want to make them available to R. To do that we have a few routes.

### Registering C++ functions outside of a package context

After writing our hello world program in foo.cpp we can use `cpp_source()` to compile and register the function to R.

``` r
cpp_source(file = "src/foo.cpp")
```

Now the function is available in R

```{r}
hello_world()
```

Similarly we can use the helper `cpp_eval` to run simple expressions and return the result without needing to include cppally.hpp and register the function.

```{r}
cpp_eval('print("Hello World Again!")')
```

**Note** - For the rest of the examples it is assumed that the following code is always included beforehand.

``` cpp
#include <cppally.hpp>
using namespace cppally;
```

### Registering C++ functions inside a cppally-linked package

Since cppally is header-only, we can include the headers directly into our own package.

### General steps to using cppally in a package

1.  Create package (if you haven't already done so) using `usethis::create_tidy_package()`
2.  Run `cppally::use_cppally()`
3.  Run `cppally::document()`

This will automatically add the necessary package content needed to start working with cppally. For continuous development, use `cppally::load_all()` to compile and register cppally tagged functions, including our hello world function.

**Note:** We aim to integrate cppally registration into the `devtools` framework for ease-of-use.

## C++ types

cppally offers a rich set of R types in C++ that are NA-aware. This means that common arithmetic and logical operations will account for `NA` in a similar fashion to R.

## Scalars

### logical scalar - `r_lgl`

cppally's scalar version of `logical`, `r_lgl` can represent true, false or NA.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(paste(
  single_exprs[["r_true_val"]],
  single_exprs[["r_false_val"]],
  single_exprs[["r_na_val"]],
  sep = "\n"
))
```

```{r, echo=FALSE}
r_true_val()
r_false_val()
r_na_val()
```

Logical operators work just like in R

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["lgl_ops"]])
```

```{r}
lgl_ops()
```

**Using `r_lgl`** in if-statements

For type-safety reasons `r_lgl` cannot be implicitly converted to `bool` except in if-statements where an error is thrown if the value is `NA`.

**DON'T** do this:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["bad_lgl_print"]])
```

```{r, error=TRUE}
bad_lgl_print(TRUE)
bad_lgl_print(FALSE)
bad_lgl_print(NA) # Can't implicitly convert NA to bool
```

**DO** this:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["good_lgl_print"]])
```

```{r}
good_lgl_print(TRUE)
good_lgl_print(FALSE)
good_lgl_print(NA) # NA is handled explicitly so no issues
```

We can also use `r_lgl` members `is_true()` and `is_false()` which return `bool` and are equivalent to R's `isTRUE()` and `isFALSE()`

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["also_good_lgl_print"]])
```

```{r}
also_good_lgl_print(TRUE)
also_good_lgl_print(FALSE)
also_good_lgl_print(NA) # Falls into 'not true' branch here as expected
```

**Important:** The `&&` and `||` operators for `r_lgl` do **NOT short-circuit** like they do for `bool`. Both sides of the expression are always evaluated. If you specifically require short-circuiting behaviour, use `is_true()` and `is_false()` before using `&&` and `||`.

All cppally scalar types are implemented as structs that contain the underlying C/C++ types as well as other member functions.

| cppally type | Description                 | Built on               |
|:-------------|-----------------------------|:-----------------------|
| `r_lgl`      | Scalar logical              | `int`                  |
| `r_int`      | Scalar integer              | `int`                  |
| `r_int64`    | Scalar 64-bit integer       | `int64_t`              |
| `r_dbl`      | Scalar double               | `double`               |
| `r_str`      | Scalar string               | `r_sexp`               |
| `r_str_view` | Scalar string (view)        | `SEXP`                 |
| `r_cplx`     | Scalar double complex       | `std::complex<double>` |
| `r_raw`      | Scalar raw                  | `unsigned char`        |
| `r_sym`      | Symbol                      | `SEXP`                 |
| `r_date`     | Scalar date                 | `r_dbl`                |
| `r_psxct`    | Scalar date-time            | `r_dbl`                |
| `r_sexp`     | Generic R object (SEXP)[^1] | `SEXP`                 |

[^1]: `r_sexp` represents a generic R object which can include cppally vectors. We will explain how to disambiguate `r_sexp` later which is most useful when working with lists and data frames

### NA values

Use `is_na()` to check that a value is NA and `na<T>()` to generate NA values.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_na_check"]])
```

```{r, echo=FALSE}
r_na_check()
```

### C++ NA values and their R C API equivalents

| Type                     | Value                   | R C API Value  |
|--------------------------|-------------------------|----------------|
| `r_lgl`                  | `na<r_lgl>()`/`r_na`    | `NA_LOGICAL`   |
| `r_int`                  | `na<r_int>()`           | `NA_INTEGER`   |
| `r_int64`                | `na<r_int64>()`         | Not applicable |
| `r_dbl`                  | `na<r_dbl>()`           | `NA_REAL`      |
| `r_str`                  | `na<r_str>()`           | `NA_STRING`    |
| `r_cplx`                 | `na<r_cplx>()`          | Not applicable |
| `r_sym`                  | Not applicable          | Not applicable |
| `r_sexp`[^2]<br><br><br> | `na<r_sexp>()`/`r_null` | `R_NilValue`   |

[^2]: Having an `NA` sentinel for `r_sexp` is very useful when writing templates involving vectors. For this reason the `NA` sentinel is `r_null`. This doesn't mean `is_na(r_null)` is true, and is intentionally not true because it is not a scalar and therefore cannot be `NA`. As `r_null` represents the absence of a tangible R object, it can be thought of as a zero-length object and since all `NA` values are represented as length-1 vectors (in R), `is_na(r_null)` should not return true.

## Checking equality

There are two ways to check for exact equality of cppally scalars - with the `==` operator or with `identical()`.

The cppally `==` operator always returns `r_lgl` and `identical()` always returns `bool`, which is a particularly important distinction when dealing with `NA` values as the former can represent `NA` while the latter cannot.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cppally_equality"]])
```

```{r}
cppally_equality()
```

`identical()` can not only compare scalars, but also vectors, lists, factors, and data frames.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cppally_identical"]])
```

```{r}
cpp_identical(3L, 3L)
cpp_identical(NA, NA)
cpp_identical(3L, 3) # int != double
cpp_identical(1:10, 1:10)
cpp_identical(list(1, 2, 3), list(3, 2, 1))
cpp_identical(iris, iris)
```

## Scalar operators

cppally also defines arithmetic and relational comparison operators for its scalar types. Like the logical and equality operators seen earlier, they are all `NA`-aware.

### Scalar arithmetic operators

**Addition**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_plus"]])
```

```{r, echo=FALSE}
r_plus()
```

**Subtraction**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_minus"]])
```

```{r, echo=FALSE}
r_minus()
```

**Multiplication**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_mult"]])
```

```{r, echo=FALSE}
r_mult()
```

**Division**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_div"]])
```

```{r, echo=FALSE}
r_div()
```

**Addition with `NA`**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_plus_na"]])
```

```{r, echo=FALSE}
r_plus_na()
```

**Subtraction with `NA`**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_minus_na"]])
```

```{r, echo=FALSE}
r_minus_na()
```

**Multiplication with `NA`**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_mult_na"]])
```

```{r, echo=FALSE}
r_mult_na()
```

**Division with `NA`**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_div_na"]])
```

```{r, echo=FALSE}
r_div_na()
```

### Scalar relational operators

**Less than**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_lt"]])
```

```{r, echo=FALSE}
r_lt()
```

**Less than or equal to**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_lte"]])
```

```{r, echo=FALSE}
r_lte()
```

**Greater than**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_gt"]])
```

```{r, echo=FALSE}
r_gt()
```

**Greater than or equal to**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_gte"]])
```

```{r, echo=FALSE}
r_gte()
```

**Less than with `NA`**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_lt_na"]])
```

```{r, echo=FALSE}
r_lt_na()
```

**Less than or equal to with `NA`**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_lte_na"]])
```

```{r, echo=FALSE}
r_lte_na()
```

**Greater than with `NA`**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_gt_na"]])
```

```{r, echo=FALSE}
r_gt_na()
```

**Greater than or equal to with `NA`**

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_gte_na"]])
```

```{r, echo=FALSE}
r_gte_na()
```

Other defined operators not showcased: `++`, `--`, `+=`, `-=`, `*=`, `/=`, `%=`, `-`, `%`, `|`, `&`, `!`

## Vectors

cppally vectors are templated and can be thought of as containers of scalar elements like `r_int`, `r_dbl`, etc.

We can create vectors like so

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["new_integer_vector"]])
```

```{r}
new_integer_vector(3)
```

### inline vectors

To create inline vectors, use `make_vec<>`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["make_vec_dbl"]])
```

```{r, echo=FALSE}
make_vec_dbl()
```

We can add names on the fly with `arg()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["make_vec_dbl_named"]])
```

```{r, echo=FALSE}
make_vec_dbl_named()
```

In R a list is a generic vector, so cppally defines lists as `r_vector<r_sexp>`, a vector of the generic type `r_sexp`.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["make_vec_sexp"]])
```

```{r, echo=FALSE}
make_vec_sexp()
```

A list of all cppally vectors of length 0

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["all_vectors"]])
```

```{r}
all_vectors()
```

## Scalar math

There is a rich suite of math functions that accept cppally types.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cppally_math"]])
```

```{r}
cppally_math(2.5)
cppally_math(NA)
```

## Coercion

To coerce from one scalar to another we can use `as<T>`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["double_to_int"]])
```

```{r}
double_to_int(pi)
double_to_int(NA_real_)
```

We can also coerce from one vector type to another

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["to_int_vec"]])
```

```{r}
to_int_vec(c(0, 1.5, NA))
```

Since `as<T>` is extremely flexible, we can also coerce from a scalar to a vector or vice versa

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["coercions"]])
```

```{r}
coercions()
```

We can even coerce to and from C++ vectors

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["to_from_cpp_vec"]])
```

```{r}
cpp_vectors_example(41L)
```

While coercing to a `std::vector` just to push back an element before coercing back might not be the most efficient, it does showcase how easy it is to work with cppally vectors and C++ vectors.

## Strings

cppally provides the useful string type `r_str`

We can create R strings easily

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_str_hello"]])
```

```{r, echo=FALSE}
r_str_hello()
```

To get a C or C++ string, use the members `c_str()` and `cpp_str()` respectively

C string via `c_str()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_str_hello_c_str"]])
```

```{r, echo=FALSE}
r_str_hello_c_str()
```

C++ string_view via `cpp_str()`

This can be converted into a std::string via its constructor

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["str_concatenate"]])
```

```{r}
str_concatenate("hello", "how are you?", sep = ", ")
```

## Dates and date-times

cppally offers two classes: `r_date` for dates and `r_psxct` for date-times (based on POSIX time). These classes offer a rich set of member functions to perform date arithmetic and manipulation.

To quickly re-familiarise ourselves - R Dates and Date-times are essentially double-valued offsets from Unix time. Unix time is defined as the number of seconds since 1 Jan 1970, and so `r_date` is effectively an offset representing the number of days since epoch, and similarly `r_psxct` is an offset representing the number of seconds since epoch.

To create a new `r_date`, use the `r_date(year, month, day)` constructor.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["create_date"]])
```


```{r}
create_date(1970, 1, 1); # This builds the Unix epoch date
```

You can also build an `r_date` from an epoch offset.

```{r}
date_from_epoch_offset(0) # Also builds the Unix epoch
```

Similarly, to create a new `r_psxct`, use the `r_psxct(year, month, day, hour, minute, second)` constructor.

```{r}
create_datetime(1970, 1, 1, 0, 0, 0); # Unix epoch date-time
datetime_from_epoch_offset(0)         # Unix epoch from an offset
```

To get the current date and date-time, use `r_date::today()` and `r_psxct::now()` respectively.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["get_time"]])
```

```{r}
get_today()
get_now()
```

**Note:** `r_psxct` currently only supports UTC and no other time-zones.

#### Adding days, months, and years

`r_date` and `r_psxct` both have the template function `add<>`, which allows us to add years, months, weeks, days, hours, minutes and seconds to our dates and date-times.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["add_time"]])
```

```{r}
y2k <- create_date(2000, 1, 1)

# Day after y2k
y2k |>
  add_days(1)

# Week after y2k
y2k |>
  add_days(7)

# Month after y2k
y2k |>
  add_months(1)

# Year after y2k
y2k |>
  add_months(12)
```

### Accessing date components

Those familiar with lubridate will instantly be familiar with the naming of these functions, something that was an intentional design choice to reduce confusion for those who are familiar with tidyverse.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["get_time_components"]])
```

```{r}
get_year(y2k)
get_month(y2k)
get_week(y2k)
get_day(y2k) # Day of the month
get_wday(y2k, week_start = 1) # week_start = 1 = Mon
get_yday(y2k)
```

We can even retrieve the ISO week and ISO year.

```{r}
get_iso_week(y2k) # Last (52nd) ISO week of the ISO year
get_iso_year(y2k) # ISO year is still 1999
```

### Date rounding

We can also round dates using the member functions `floor()`, `ceiling()`, and `round()`.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["time_round"]])
```

```{r}
# Start of the month after advancing 100 days
y2k |> 
  add_days(100) |> 
  floor_month()

# ceiling() does not switch on boundary
y2k |> 
  ceiling_year()

y2k |> 
  add_days(1) |> 
  ceiling_year()

y2k |> 
  round_week(week_start = 1)
```

### Difference between two dates

To calculate the time difference between two dates or date-times, use `time_diff<>`.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["time_diff"]])
```

```{r}
x <- y2k
y <- x |> add_months(3)

diff_days(x, y)
diff_months(x, y)
```

### Fractional months

Both `add<"months">` and `time_diff<"months">` can deal with fractional months. 

The previous sections dealt with dates, so we need to define functions that work with `r_psxct` (date-times).

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["datetime_months"]])
```

```{r}
y2k_datetime <- as_dt(y2k)

mid_month <- y2k_datetime |> 
  dt_add_months(0.5)

# Half-way through the month 
mid_month
```

The difference between the start of the month and halfway through the month 
should return 0.5, confirming that fractional month arithmetic is working.

```{r}
dt_diff_months(y2k_datetime, mid_month)
```


## Symbols

Symbols have class `r_sym` and can be created directly from a string literal

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_sym_new"]])
```

```{r, echo=FALSE}
r_sym_new()
```

Or from a cppally string

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_sym_from_str"]])
```

```{r, echo=FALSE}
r_sym_from_str()
```

## Cached strings & symbols

cppally provides an efficient caching strategy for constructing cppally strings/symbols from string literals

`cached_str<>`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["cached_str_demo"]])
```

```{r, echo=FALSE}
cached_str_demo()
```

This initialises the string once, caches it (to R's CHARSXP pool), and efficiently re-uses the cached string for each subsequent call.

We can cache symbols in a similar way

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["cached_sym_demo"]])
```

```{r, echo=FALSE}
cached_sym_demo()
```

## Lists

`r_sexp` is generally interpreted as an "element of a list" since lists are defined as `r_vector<r_sexp>`, a vector that holds generic `r_sexp` elements.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["new_list"]])
```

```{r}
new_list(0)
new_list(3)
```

The problem with a class like `r_sexp` is that it is by design generic and therefore difficult to work with in C++. To disambiguate the actual type we can use `r_sexp_visit()` via a C++ lambda.

**Example:** using `r_sexp_visit()` to resize every vector to length n in-place

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["resize_all"]])
```

## Factors

We can create a factor via `r_factors()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["new_factor"]])
```

```{r}
new_factor(letters)
```

In cppally, like R, factors are not vectors and therefore do not satisfy the RVector concept. To access the underlying integer codes vector, use the public `codes()` member function

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["factor_codes"]])
```

```{r}
letter_fct <- new_factor(letters)

letter_fct |>
    factor_codes()
```

## Calling R functions

cppally provides the `r_function` class, which can be used to easily call any R function.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["r_get_today"]])
```

```{r}
r_get_today()
```

To get a function from a specific package, use `pkg_env`, a helper that returns the namespace of a package. The result of `pkg_env` can be passed to the `env` (or second) arg of `r_function`'s symbol and string constructors.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["base_sum_fn"]])
```

```{r}
base_sum_fn()
```

`r_function` also allows named arguments to be passed.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["base_sum"]])
```

```{r}
base_sum(c(NA, 1:3, NA), na_rm = TRUE)
```

## Value matching

Use `r_vector` member `find()` to find the **0-indexed** locations of a scalar value.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["find_empty_string"]])
```

```{r}
x <- c("zero", "one", "two", "three", "four")

find_empty_string(x)

# Add empty strings
x[c(1, 3)] <- ""

find_empty_string(x)
```

To find the locations of multiple values (first match), use `match()`. It works like R's `base::match()` but is 0-indexed.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["match_strs"]])
```

```{r}
letters

vowels <- c("a", "e", "i", "o", "u")

# cppally::match is 0-indexed
match_strs(vowels, letters)
```

cppally provides the `IS_IN` infix operator, identical to R's `%in%`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cpp_in"]])
```

```{r}
cpp_in(c("a", "A", NA), letters)
```

To mimic R's new `%notin%` operator, simply use the logical negation operator.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cpp_not_in"]])
```

**Technical note:** cppally internally negates the intermediate result of `x IS_IN table` in-place in this particular case because it satisfies specific properties of exclusivity which while not covered here, may be covered in a later vignette. This in-place negation is naturally efficient as it avoids allocating a new vector.

```{r}
cpp_not_in(c("a", "A", NA), letters)
```

## Subsetting

Subsetting is 0-indexed in cppally, as is all other indexing.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cpp_subset"]])
```

Subsetting with **integer** indices

```{r}
x <- 1:10
cpp_subset(x, 0L) # index 0 is 1st value
cpp_subset(x, 9L) # index 9 is last value here
cpp_subset(x, 10L) # index 10 is out-of-bounds so NA is returned
cpp_subset(x, NA_integer_) # NA is returned with integer NA index
```

Subsetting with **logical** indices

```{r}
cpp_subset(x, x > 5)
cpp_subset(x, x > 100)

# It differs to base subsetting (via `[`)
cpp_subset(x, rep(NA, 10)) # cppally only returns values associated with TRUE
x[rep(NA, 10)] # base R returns NA values when subsetting with NA logicals
```

base R performs negative subsetting by using negative numbers. This is not possible with our 0-indexed subset as we would have to also represent negative zero to subset everything except the first element at index 0. Instead cppally has an `invert` argument.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cpp_negative_subset"]])
```

```{r}
cpp_negative_subset(x, 0L) # Everything but 1st
cpp_negative_subset(x, x > 5) # Everything except where x > 5
cpp_negative_subset(x, integer()) # Everything
```

Named subsetting is also supported. cppally internally hashes the vector names and performs hash lookups. For more info, see the 'Vector Names Hashing' vignette.

```{r}
names(x) <- LETTERS[seq_along(x)]

cpp_subset(x, c("A", "J", "Z"))
cpp_negative_subset(x, "A")
```

## Concepts and Templates

One of the most powerful features of C++20 are concepts. These allow users to write human-readable templates and constraints.

When writing your own templates, it is necessary to place them in headers for cppally registration to work correctly.

Let's practice by creating the `abs()` function in C++ using templates and the `RNumber` concept.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cpp_abs"]])
```

What's nice is that it works correctly for integers and doubles while simultaneously preserving their type

```{r}
cpp_abs(-4.2)
cpp_abs(-3L)

class(cpp_abs(-4.2)) # Double preserved
class(cpp_abs(-3L)) # Integer preserved
```

This type of programming is historically tricky within the R C API and typically necessitates a switch statement that switches on the object's type, handling each type separately. With our `abs()` template, the logic is correctly handled with one set of operations.

### How it works

The top-line `template <RNumber T>` declares a template that encapsulates `T`, an `RNumber` - a concept that contains `r_int`, `r_int64` and `r_dbl`

If x is NA then we immediately also return NA via `na<T>()` which is a templated function that returns NA of the input type `T`.

To correctly register templates, the '[[cppally::register]]' tag must always go above the function name.

``` cpp
template <typename T>
[[cppally::register]] // <--- Here
T foo(T x){
  return x;
}
```

### Templates without function arguments

Explicit instantiation (from R) is unfortunately not possible and template types must be deduced from supplied arguments.

``` cpp
template <typename T>
[[cppally::register]]
T foo(){
    return T();
}
```

Here `foo()` will not be compiled because the function has no arguments that let the compiler automatically deduce what `T` is. In C++ you would always call this function like so: `foo<T>()`. Unfortunately we can't do that from R directly.

You may get a cryptic compiler error like this

``` cpp
error: no matching function for call to 'foo()'
[]<typename T>() -> decltype(cpp_to_r(::foo())) {
```

along with an equally cryptic note

``` cpp
note:   couldn't deduce template parameter 'T'
[]<typename T>() -> decltype(cpp_to_r(::foo())) {
```

Even though these kinds of templates can be written with cppally in C++, they cannot be exported to R.

An obvious and somewhat ugly workaround is to include a prototype argument that allows the template parameter to be deduced from.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["scalar_default"]])
```

```{r}
scalar_default(integer(1)) # Default is 0L
scalar_default(numeric(1)) # Default is 0.0
scalar_default(character(1)) # Default is ""
```

Exporting variadic templates are also not supported. The best alternative is to use lists (`r_vector<r_sexp>`).

In the above example we used the `RScalar` concept which includes all cppally scalar types (excluding `r_sexp`). For a list of all cppally concepts, please see the **Annex**

## Attributes

Attributes can be manipulated via functions defined in the attr namespace.

**Example:** Adding names to a list

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["setnames"]])
```

```{r}
set.seed(42)
norm_samples <- lapply(1:5, \(x) rnorm(10, mean = x))
set_list_names(norm_samples, paste0("sample_", 1:5))
```

More useful attribute helpers

- `get_attrs()` - Returns a list of attributes (possibly `r_vector<r_sexp>(r_null)`)
- `set_attrs()` - Sets attributes to ones specified. Note: replaces any current attributes
- `clear_attrs()` - Removes all attributes
- `set_attr()` - Set a single attribute
- `get_attr()` - Get a single attribute
- `inherits1()` - Does object inherit class?
- `inherits_any()` - Does object inherit at least one of the specified classes?
- `inherits_all()` - Does object inherit all of the specified classes?
- `modify_attrs()` - Modifies current attributes but doesn't remove any existing ones

## Regular sequences

There are two core functions for generating regular sequences - `sequence()` and `seq()`. `seq()` behaves exactly like the R equivalent `base::seq()`, and `sequence()` behaves like the R equivalent `base::sequence()`, with the exception that it accepts scalar arguments instead of vector ones, and also works with decimal increments.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["cpp_seq"]])
```

```{r, echo=FALSE}
cpp_seq()
```

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["cpp_sequence"]])
```

```{r, echo=FALSE}
cpp_sequence()
```

We can also use `sequence()` to easily replicate `base::seq_len()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cpp_seq_len"]])
```

```{r}
cpp_seq_len(5)
```

It is also straightforward to replicate `base::sequence()` with `pmap()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cpp_sequences"]])
```

```{r}
cpp_sequences(1:3, from = 0L, by = 1L) |> 
  unlist()

# Same as base R
sequence(1:3, from = 0L, by = 1L)
```

## Using multiple OpenMP threads

To set the number of global OpenMP threads, use `cppally::set_threads()`.

```cpp
cppally::set_threads(4); // Sets the number of threads to 4
```

To see the number of currently available threads, use `cppally::get_threads()`.

From R we can then register these C++ function and use them.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(benchmark_examples[["cpp_set_threads"]])
```

```{r}

cpp_set_threads(2) # Set to 2 threads
cpp_get_threads()  # Number of threads now available

cpp_set_threads(1) # Reset back to 1
cpp_get_threads()  # Should return 1
```

Because this is unique to each dll file, setting threads in one R package doesn't affect another. Once threads have been set, all cppally code that 
can make use of them, will use them. This means you should speed improvements across many cppally functions.

## Stats functions

cppally provides some stats functions which are both common and difficult to implement efficiently and safely.

- `sum()` - Sum of values
- `mean()` - Average of values
- `range()` - Min and max range of values
- `var()` - Variance


```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(benchmark_examples[["cpp_stats"]])
```

```{r}
library(bench)

set.seed(1)
x <- sample.int(10, 5e05, replace = TRUE)
x[sample.int(10^5, 10^4)] <- NA # Add many NAs

# Sum (ignoring NAs)
mark(
  cppally_sum = cpp_sum(x, na_rm = TRUE),
  base_sum = sum(x, na.rm = TRUE)
)

# Sum (not ignoring NAs)
mark(
  cppally_sum = cpp_sum(x, na_rm = FALSE),
  base_sum = sum(x, na.rm = FALSE)
)

# Range (ignoring NAs)
mark(
  cppally_range = cpp_range(x, na_rm = TRUE),
  base_range = range(x, na.rm = TRUE)
)

# Range (not ignoring NAs)
mark(
  cppally_range = cpp_range(x, na_rm = FALSE),
  base_range = range(x, na.rm = FALSE)
)

# Variance (ignoring NAs)
mark(
  cppally_var = cpp_var(x, na_rm = TRUE),
  base_var = var(x, na.rm = TRUE)
)

# Variance (not ignoring NAs)
mark(
  cppally_var = cpp_var(x, na_rm = FALSE),
  base_var = var(x, na.rm = FALSE)
)

# Using multiple threads

cpp_set_threads(4)

# Sum (ignoring NAs)
mark(
  cppally_sum = cpp_sum(x, na_rm = TRUE),
  base_sum = sum(x, na.rm = TRUE)
)

# Sum (not ignoring NAs)
mark(
  cppally_sum = cpp_sum(x, na_rm = FALSE),
  base_sum = sum(x, na.rm = FALSE)
)

# Range (ignoring NAs)
mark(
  cppally_range = cpp_range(x, na_rm = TRUE),
  base_range = range(x, na.rm = TRUE)
)

# Range (not ignoring NAs)
mark(
  cppally_range = cpp_range(x, na_rm = FALSE),
  base_range = range(x, na.rm = FALSE)
)

# Variance (ignoring NAs)
mark(
  cppally_var = cpp_var(x, na_rm = TRUE),
  base_var = var(x, na.rm = TRUE)
)

# Variance (not ignoring NAs)
mark(
  cppally_var = cpp_var(x, na_rm = FALSE),
  base_var = var(x, na.rm = FALSE)
)

# Reset to single-threaded
cpp_set_threads(1)

```

### A note on cppally::sum

`cppally::sum()` is written to use SIMD instructions for performance where possible. It also uses higher width integers for accuracy when dealing with integer vectors. For example, if summing a 32-bit integer vector, it uses intermediate 64-bit integers, and likewise if summing a 64-bit integer vector, it uses intermediate 128-bit integers (if your compiler supports it). Doing it this way avoids integer overflow entirely for vectors whose length is less than 2^32. It also avoids intermediate integer overflow that would have arisen if same-width integers were used. For example, 
for `x <- .Machine$integer.max * c(-1L, -1L, 1L, 1L)`, the sum of x is trivially 0, but the intermediate addition of `x[1]` and `x[2]` would result in overflow if one were to use 32-bit integers. 

## Unique values

`n_unique()` is a convenient helper to calculate the number of unique values. `unique()` is also highly performant.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(benchmark_examples[["cpp_unique"]])
```

```{r}
x <- sample(1:100, 10^5, replace = TRUE)

mark(
  base_n_unique = length(unique(x)),
  cppally_n_unique = cpp_n_unique(x)
)

mark(
  base_unique = unique(x),
  cppally_unique = cpp_unique(x, sort = FALSE)
)

mark(
  base_sorted_unique = sort(unique(x)),
  cppally_sorted_unique = cpp_unique(x, sort = TRUE)
)
```

## Sorting

cppally also offers `order()` and `sort()` for sorting purposes. A hybrid 
approach is used, incorporating various sorting algorithms, including 
ska sort (a variant on radix sort), counting sort, quick sort, etc.
These have been implemented with speed being a key focus.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(benchmark_examples[["cpp_sort"]])
```

`cppally::sort` always sorts `NA` values last and so is equivalent to 
`base::sort(..., method = "radix", na.last = TRUE)`.

```{r}
r_radix_sort <- function(x){
  sort(x, method = "radix", na.last = TRUE)
}
r_radix_order <- function(x){
  order(x, method = "radix", na.last = TRUE)
}

# Small data
mark(cpp_sort(c(3, 2, 1)), r_radix_sort(c(3, 2, 1)))
mark(cpp_order(c(3, 2, 1)), r_radix_order(c(3, 2, 1)))

# Integer data
ints <- sample.int(20, 2e05, replace = TRUE)
mark(cpp_sort(ints), r_radix_sort(ints))
mark(cpp_order(ints), r_radix_order(ints))

# Double-floating point precision data
dbls <- as.double(ints)
mark(cpp_sort(dbls), r_radix_sort(dbls))
mark(cpp_order(dbls), r_radix_order(dbls))

# Floating point data with decimals
dcmls <- rnorm(length(dbls))
mark(cpp_sort(dcmls), r_radix_sort(dcmls))
mark(cpp_order(dcmls), r_radix_order(dcmls))

# String data
strs <- round(dcmls, 2) |> paste0()
mark(cpp_sort(strs), r_radix_sort(strs))
mark(cpp_order(strs), r_radix_order(strs))
```

## Grouping

Some areas of analysis require working with groups, with sometimes very large numbers of them. cppally provides a powerful class `groups`, along with `make_groups()`, a function that creates `groups` from vectors.

The structure of `groups` is:

  - `r_vector<r_int> ids` - The group IDs
  - `int n_groups` - Number of unique groups
  - `bool ordered` - Do the group IDs specify a sorting order, or are they by order-of-first-appearance?
  - `bool sorted` - Are the group IDs sorted? (This can also be true for order-of-first-appearance IDs)
  - `r_vector<r_int> starts()` - Returns a vector of start locations of each unique group, signifying the location in the data at which each group initially appeared
  - `r_vector<r_int> counts()` - Returns a vector of frequency counts of each unique group
  - `r_vector<r_int> order()` - Returns a 0-indexed permutation vector that can be used to sort group IDs

**Note:** When the groups are in sorted-order (where `sorted` is true), member functions like `starts()` and `counts()` use a hybrid search strategy combining linear, gallop, and binary searching. This strategy ensures that search time complexity scales with the number of groups and not the size of the data.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(benchmark_examples[["make_groups"]])
```


```{r}
set.seed(1)
x <- sample(c("A", "B", "C"), 20, replace = TRUE)

# Convert into group info

letter_groups <- group_info(x)

letter_groups$group_id # Group IDs
letter_groups$n_groups # Number of groups
letter_groups$ordered  # Groups were collected using order-of-first-appearance
letter_groups$sorted   # Are group IDs in sorted order?
letter_groups$starts   # Data start locations (0-indexed)
letter_groups$counts   # Group counts
letter_groups$order    # Order permutation (0-indexed)
letter_groups$group_names # Names of unique groups
```

In real analysis, you normally wouldn't need all of this metadata. For example, to return a data frame of unique groups (keys) and their associated counts, we only need `starts()` and `counts()`.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(benchmark_examples[["group_counts"]])
```

```{r}
group_counts(x)
```

To illustrate how fast `make_groups()` is, let's compare cppally-derived group counts of a relatively large sample to base-R derived group counts.

```{r}
# 10k groups, 100k data points
x <- sample.int(10^4, 10^5, replace = TRUE)
mark(
  cppally_group_counts = group_counts(x)$count,
  base_table = table(factor(x, levels = unique(x))) |> as.integer()
)
```

## Annex

### Unsupported classes in R-registered templates

`r_sym` and `r_function` are both unsupported in R-registered template functions. This only applies when used as template arguments. They will work as normal where function arguments are non-templated.

Good:

```cpp
template <RStringType T>
[[cppally::register]]
r_lgl foo(r_sym x, T y){
    return as<r_str>(x) == y;
}
```

Bad: 

```cpp
template <RStringType T, typename Symbol>
requires (is<Symbol, r_sym>)
[[cppally::register]]
r_lgl foo(Symbol x, T y){
    return as<r_str>(x) == y;
}
```

### All core cppally concepts

- RNumber - Includes `r_int`, `r_int64` and `r_dbl`

- RIntegerType - Includes `r_lgl`, `r_int`, `r_int64`

- RIntegerNumber - Includes `r_int` and `r_int64`

- RMathType - Includes `r_lgl`, `r_int`, `r_int64` and `r_dbl`

- RStringType - Includes `r_str` and `r_str_view`

- RScalar - Includes `r_lgl`, `r_int`, `r_int64`, `r_dbl`, `r_str`, `r_str_view`, `r_cplx`, `r_raw`, `r_date` and `r_psxct`

- RVal - Includes anything a cppally vector (`r_vector<>`) can contain: RScalar +`r_sexp`

- RVector - Includes `r_vector<T>` where `T` is an RVal

- RFactor - Factors

- RDataFrame - Data frames

- RComposite - Includes vectors, factors and data frames

- RTimeType - Includes `r_date` and `r_psxct`

- RNumericType - Numeric types, including RMathType and RTimeType

- RSortableType - Includes RNumericType and RStringType (strings can also be sorted)

- RAtomicVector - A vector that contains RScalar elements

- CppallyType - Any R type defined by R, including RVal, RVector, RFactor, RDataFrame, RSymbol

- CppType - Anything that is not an CppallyType

- CastableToRScalar - Anything that can be constructed or cast into an RScalar (which also includes RScalar)

Other useful type traits

- `unwrap_t` - Returns the underlying unwrapped type
- `as_r_scalar_t` - Returns the equivalent RScalar type
- `as_r_composite_t` - Returns the equivalent RComposite type
- `common_r_t` - Returns the common cppally type between 2 types

### Accessing the underlying types and values

While it is generally not recommended to access the underlying values, you can do so with `unwrap()` which returns the underlying C/C++ value. For example, `unwrap(r_int(5))` will return an `int` of value `5`.

To access the underlying type, use `unwrap_t<>` which always aligns with `unwrap()`

```cpp
int one = unwrap(r_int(1));
double two = unwrap(r_dbl(2));

// Confirming that unwrap_t produces the required types as well
static_assert(std::is_same_v<int, unwrap_t<r_int>>);
static_assert(std::is_same_v<double, unwrap_t<r_dbl>>);

// unwrap_t<r_str> = SEXP
// cppally recommends against using BOTH the cppally API and the R C API together.
unwrap_t<r_str> r_c_api_blank_str = unwrap(r_str(""));
```
