
---
title: "Getting started with cppally"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cppally}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

```{r, include = FALSE}
required <- c("bench", "brio", "callr", "cli", "decor",
              "desc", "glue", "purrr", "readr", "stringr",
              "utils", "vctrs", "withr")
if (!all(vapply(required, requireNamespace, logical(1), quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
  knitr::knit_exit()
}
```


Let's briefly show some of the capabilities of cppally, from its custom
C++ scalar and vectors, to using templates and concepts.

## Setup

Let's start by loading cppally

```{r}
library(cppally)
```

```{r, include=FALSE}
# Helpers to compile all examples in debug mode
cpp_source <- function(..., code, debug = TRUE, env = parent.frame()){
  preamble <- c("#include <cppally.hpp>", "using namespace cppally;")
  code <- paste(c(preamble, code), collapse = "\n")
  cppally::cpp_source(debug = debug, env = env, code = code, ...)
}
cpp_eval <- function(..., debug = TRUE, env = parent.frame()){
    cppally::cpp_eval(debug = debug, env = env, ...)
}

# Helpers to source and display C++/R code
chunk_impl <- function(x, language){
  paste0("```", language, "\n", x, "\n```\n")
}
as_code_chunk <- function(x, language){
  cat(chunk_impl(x, language))
}
as_cpp_chunk <- function(x){
  as_code_chunk(x, "cpp")
}

# Pre-register named single-line expressions so they can be referenced later
register_single_exprs <- function(exprs, env = parent.frame(), ...){
  if (is.null(names(exprs))){
    stop("`exprs` must be named")
  }
  utils::getFromNamespace("source_single_exprs", "cppally")(
    exprs, env = env, ...
    )
  wrappers <- setNames(
    lapply(seq_along(exprs), \(i) {
      fn <- get(paste0("f", i), envir = env)
      function() {
        out <- fn()
        if (out[["is_void"]]) invisible() else out$result
      }
    }),
    names(exprs)
  )
  list2env(wrappers, envir = env)
  invisible()
}
```

```{r, include=FALSE}
# Compile necessary examples in one-go
# as it's faster when building the vignette

examples <- c(
  hello_world = '
[[cppally::register]]
void hello_world(){
  print("Hello World!");
}',
  lgl_ops = '
[[cppally::register]]
r_vec<r_lgl> lgl_ops(){
  return make_vec<r_lgl>(
    r_true || r_false, // true
    r_true && r_false, // false
    r_na || r_true,    // true
    r_na && r_true,    // NA
    r_na && r_false,   // false
    r_na || r_na,      // NA
    r_na && r_na      // NA
  );
}
',
  bad_lgl_print = '
[[cppally::register]]
void bad_lgl_print(r_lgl condition){
  if (condition){
    print("true");
  } else {
    print("false");
  }
}
',
  good_lgl_print = '
[[cppally::register]]
void good_lgl_print(r_lgl condition){
  if (is_na(condition)){
    print("NA");
  } else if (condition){
    print("true");
  } else {
    print("false");
  }
}
',
  also_good_lgl_print = '
[[cppally::register]]
void also_good_lgl_print(r_lgl condition){
  if (condition.is_true()){
    print("true");
  } else {
    print("not true");
  }
}
',
  new_integer_vector = '
// Integer vector of size n
[[cppally::register]]
r_vec<r_int> new_integer_vector(int n){
  r_vec<r_int> int_vctr(n, /*fill = */ r_int(0));
  return int_vctr;
}
',
  all_vectors = '
[[cppally::register]]
r_vec<r_sexp> all_vectors(){
  return make_vec<r_sexp>(
    arg("logical") = r_vec<r_lgl>(),
    arg("integer") = r_vec<r_int>(),
    arg("integer64") = r_vec<r_int64>(), // Requires bit64
    arg("double") = r_vec<r_dbl>(),
    arg("character") = r_vec<r_str>(),
    arg("character") = r_vec<r_str_view>(),
    arg("raw") = r_vec<r_raw>(),
    arg("date") = r_vec<r_date>(),
    arg("date-time") = r_vec<r_psxct>(),
    arg("list") = r_vec<r_sexp>()
  );
}
',
  cpp_abs = '
template <RMathType T>
[[cppally::register]]
T cpp_abs(T x){
  if (is_na(x)) return na<T>();

  if (x < 0){
    return -x;
  } else {
    return x;
  }
}
',
  scalar_default = '
// Return the default constructor result of RScalar types

template <RScalar T>
[[cppally::register]]
T scalar_default(T ptype){
	return T();
}
',
  double_to_int = '
[[cppally::register]]
r_int double_to_int(r_dbl x){
  return as<r_int>(x);
}
',
  to_int_vec = '
[[cppally::register]]
r_vec<r_int> to_int_vec(r_vec<r_dbl> x){
  return as<r_vec<r_int>>(x);
}
',
  coercions = '
[[cppally::register]]
r_vec<r_sexp> coercions(){
    r_dbl a(4.2);
    r_vec<r_dbl> b = make_vec<r_dbl>(2.5);
    return make_vec<r_sexp>(
        as<r_vec<r_int>>(a),
        as<r_int>(a),
        as<r_int>(b),
        as<r_dbl>(b)
    );
}
',
  str_concatenate = '
[[cppally::register]]
r_str str_concatenate(r_str x, r_str y, r_str sep){
  std::string left = std::string(x.cpp_str());
  std::string right = std::string(y.cpp_str());
  std::string middle = std::string(sep.cpp_str());
  std::string combined = left + middle + right;
  return r_str(combined.c_str());
}
',
  new_list = '
using list = r_vec<r_sexp>;

[[cppally::register]]
list new_list(int n){
  return list(n);
}
',
  resize_all = '
[[cppally::register]]
r_vec<r_sexp> resize_all(r_vec<r_sexp> x, r_size_t n){
    r_size_t list_length = x.length();
    for (r_size_t i = 0; i < list_length; ++i){
        visit_vector(x.view(i), [&](auto vec) {
            x.set(i, vec.resize(n));
        });
    }
    return x;
}
',
  resize_all2 = '
[[cppally::register]]
r_vec<r_sexp> resize_all2(r_vec<r_sexp> x, r_size_t n){
    r_size_t list_length = x.length();
    for (r_size_t i = 0; i < list_length; ++i){
        visit_sexp(x.view(i), [&](auto vec) {
          using vec_t = decltype(vec); // type of object `vec`
          if constexpr (RVector<vec_t>){
            x.set(i, vec.resize(n));
          } else {
            abort("Cannot resize a non-vector");
          }
        });
    }
    return x;
}
',
  new_factor = '
[[cppally::register]]
r_factors new_factor(r_vec<r_str> x){
	return r_factors(x);
}
',
  factor_codes = '
static_assert(!RVector<r_factors>);

[[cppally::register]]
r_vec<r_int> factor_codes(r_factors x){
	return x.codes();
}
',
  setnames = '
[[cppally::register]]
r_vec<r_sexp> set_list_names(r_vec<r_sexp> x, r_vec<r_str> names){
  attr::set_old_names(x, names);
  return x;
}
'
)

# Benchmarks need debug = FALSE
benchmark_examples <- c(
  cpp_n_unique = '
template <RVector T>
[[cppally::register]]
r_int cpp_n_unique(T x){
  return as<r_int>(n_unique(x));
}
',
  primitive_sum = '
[[cppally::register]]
double primitive_sum(const r_vec<r_dbl>& x){

  // r_vec<T>::data_type always returns typename T
  using data_t = typename std::remove_cvref_t<decltype(x)>::data_type;

  using primitive_t = unwrap_t<data_t>;
  primitive_t *p_x = x.data();

  r_size_t n = x.length();
  double sum = 0;

  OMP_SIMD_REDUCTION1(+:sum)
  for (r_size_t i = 0; i < n; ++i){
    sum += p_x[i];
  }
  return sum;
}
'
)

cpp_source(code = paste(examples, collapse = "\n"), debug = TRUE)
cpp_source(code = paste(benchmark_examples, collapse = "\n"), debug = FALSE)

# Single-line expressions, pre-registered as R functions of the same name.
# Each can be invoked later as e.g. `r_true_val()` to get the evaluated result.
single_exprs <- c(
  r_true_val          = 'r_true',
  r_false_val         = 'r_false',
  r_na_val            = 'r_na',
  make_vec_dbl        = 'make_vec<r_dbl>(1, 1.5, 2, na<r_dbl>())',
  make_vec_dbl_named  = '
make_vec<r_dbl>(
    arg("first") = 1,
    arg("second") = 1.5,
    arg("third") = 2,
    arg("last") = na<r_dbl>()
  )
',
  make_vec_sexp       = 'make_vec<r_sexp>(1, 2, 3)',
  r_str_hello         = 'r_str("hello")',
  r_str_hello_c_str   = 'r_str("hello").c_str()',
  r_sym_new           = 'r_sym("new_symbol")',
  r_sym_from_str      = 'r_sym(r_str("symbol_from_string"))',
  cached_str_demo     = 'cached_str<"cached_string">()',
  cached_sym_demo     = 'cached_sym<"cached_symbol">()'
)

register_single_exprs(single_exprs, debug = TRUE)
```


## Registering R functions

To make a C++ function available to R we use the `[[cppally::register]]` tag.

``` cpp
#include <cppally.hpp>
using namespace cppally;

[[cppally::register]]
void hello_world(){
  print("Hello World!");
}
```

After tagging our functions we want to make them available to R. To do that we have a few routes.

### Registering C++ functions outside of a package context

After writing our hello world program in foo.cpp we can use `cpp_source()` to compile and register the function to R.

```r
cpp_source(file = "src/foo.cpp")
```

Now the function is available in R

```{r}
hello_world()
```

Similarly we can use the helper `cpp_eval` to run simple expressions and return
the result without needing to include cppally.hpp and register the function.

```{r}
cpp_eval('print("Hello World Again!")')
```

**Note** - For the rest of the examples it is assumed that the following code
is always included beforehand.

```cpp
#include <cppally.hpp>
using namespace cppally;
```

### Registering C++ functions inside a cppally-linked package

Since cppally is header-only, we can include the headers directly into our own package.

### General steps to using cppally in a package

1. Create package (if you haven't already done so) using `usethis::create_tidy_package()`
2. Run `cppally::use_cppally()`
3. Run `cppally::document()`

This will automatically add the necessary package content needed to start working with cppally. For continuous development, use `cppally::load_all()` to compile and register cppally tagged functions, including our hello world function.

**Note:** We aim to integrate cppally registration into the `devtools` framework for ease-of-use.

## C++ types

cppally offers a rich set of R types in C++ that are NA-aware. This means that common arithmetic and logical operations will account for `NA` in a similar fashion to R.

### logical scalar - `r_lgl`

cppally's scalar version of `logical`, `r_lgl` can represent true, false or NA.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(paste(
  single_exprs[["r_true_val"]],
  single_exprs[["r_false_val"]],
  single_exprs[["r_na_val"]],
  sep = "\n"
))
```

```{r, echo=FALSE}
r_true_val()
r_false_val()
r_na_val()
```

Logical operators work just like in R

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["lgl_ops"]])
```

```{r}
lgl_ops()
```

**Using `r_lgl`** in if-statements

For type-safety reasons `r_lgl` cannot be implicitly converted to `bool` except in if-statements where an error is thrown if the value is `NA`.

**DON'T** do this:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["bad_lgl_print"]])
```

```{r, error=TRUE}
bad_lgl_print(TRUE)
bad_lgl_print(FALSE)
bad_lgl_print(NA) # Can't implicitly convert NA to bool
```

**DO** this:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["good_lgl_print"]])
```

```{r}
good_lgl_print(TRUE)
good_lgl_print(FALSE)
good_lgl_print(NA) # NA is handled explicitly so no issues
```

We can also use `r_lgl` members `is_true()` and `is_false()` which return `bool` and
are equivalent to R's `isTRUE()` and `isFALSE()`

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["also_good_lgl_print"]])
```

```{r}
also_good_lgl_print(TRUE)
also_good_lgl_print(FALSE)
also_good_lgl_print(NA) # Falls into 'not true' branch here as expected
```


All cppally scalar types are implemented as structs that contain the underlying C/C++ types as well as other member functions.

| cppally type    | Description                 | Implicitly converts to           |
| :------------ | --------------------------- | :------------------------------- |
| `r_lgl`       | Scalar logical              | `bool` **only** in if-statements |
| `r_int`       | Scalar integer              | `int`                            |
| `r_int64`     | Scalar 64-bit integer       | `int64_t`                        |
| `r_dbl`       | Scalar double               | `double`                         |
| `r_str`       | Scalar string               | `SEXP`                           |
| `r_cplx`      | Scalar double complex       | `std::complex<double>`           |
| `r_raw`       | Scalar raw                  | `unsigned char`                  |
| `r_sym`       | Symbol                      | `SEXP`                           |
| `r_date` [^1] | Scalar date                 | `double`                         |
| `r_psxct`     | Scalar date-time            | `double`                         |
| `r_sexp`      | Generic R object (SEXP)[^2] | `SEXP`                           |

[^1]: Unlike `r_str` which is composite and holds an `r_sexp` member, `r_date` and `r_psxct` instead inherit directly from `r_dbl`. This means that they can implicitly convert to `r_dbl`

[^2]: `r_sexp` represents a generic R object which can include cppally vectors. We will explain how to disambiguate `r_sexp` later which is most useful when working with lists and data frames

`NA` values can be accessed via the template function `na<T>`

### C++ NA values and their R C API equivalents

| Type                     | Value                 | R C API Value  | constexpr?[^3] |
| ------------------------ | --------------------- | -------------- | -------------- |
| `r_lgl`                  | `na<r_lgl>()`/`r_na`  | `NA_LOGICAL`   | Yes            |
| `r_int`                  | `na<r_int>()`         | `NA_INTEGER`   | Yes            |
| `r_int64`                | `na<r_int64>()`       | Not applicable | Yes            |
| `r_dbl`                  | `na<r_dbl>()`         | `NA_REAL`      | Yes            |
| `r_str`                  | `na<r_str>()`         | `NA_STRING`    | No             |
| `r_cplx`                 | `na<r_cplx>()`        | Not applicable | Yes            |
| `r_sym`                  | Not applicable        | Not applicable | No             |
| `r_sexp`[^4]<br><br><br> | `na<r_sexp>`/`r_null` | `R_NilValue`   | No             |

[^3]: In C++ constexpr is used as a keyword to declare that it's possible to evaluate values at compile-time, meaning they are known before any code is run by the user. Since `r_na` internally is the largest possible `int` which does not change and is known a priori, it is therefore a compile-time constant.
[^4]: Having an `NA` sentinel for `r_sexp` is very useful when writing templates involving vectors. For this reason the `NA` sentinel is `r_null`. This doesn't mean `is_na(r_null)` is true, and is intentionally not true because it is not a scalar and therefore cannot be `NA`.  As `r_null` represents the absence of a tangible R object, it can be thought of as a zero-length object and since all `NA` values are represented as length-1 vectors (in R), `is_na(r_null)` should not return true.


## Vectors

cppally vectors are templated and can be thought of as containers of scalar
elements like `r_int`, `r_dbl`, etc.

We can create vectors like so

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["new_integer_vector"]])
```

```{r}
new_integer_vector(3)
```

### inline vectors

To create inline vectors, use `make_vec<>`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["make_vec_dbl"]])
```

```{r, echo=FALSE}
make_vec_dbl()
```

We can add names on the fly with `arg()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["make_vec_dbl_named"]])
```

```{r, echo=FALSE}
make_vec_dbl_named()
```

In R a list is a generic vector, so cppally defines lists as `r_vec<r_sexp>`, a vector of the generic type `r_sexp`.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["make_vec_sexp"]])
```

```{r, echo=FALSE}
make_vec_sexp()
```

A list of all cppally vectors of length 0

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["all_vectors"]])
```

```{r}
all_vectors()
```

## Concepts and Templates

One of the most powerful features of C++20 are concepts. These allow users to write human-readable templates and constraints.

When writing your own templates, it is highly encouraged to place them in headers
for cppally registration to work correctly.

Let's practice by creating an absolute function in C++ using templates
and the `RMathType` concept.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["cpp_abs"]])
```

Works correctly for doubles

```{r}
cpp_abs(-5)
cpp_abs(0)
cpp_abs(100)
cpp_abs(NA_real_)
```

It also works for integers

```{r}
cpp_abs(-3L)
cpp_abs(NA_integer_)
```

The top-line `template <RMathType T>` declares a template that encapsulates `T`, an `RMathType` - a concept that contains `r_lgl`, `r_int`, `r_int64` and `r_dbl`

If x is NA then we immediately also return NA via `na<T>()` which is a templated
function that returns NA of the input type `T`.

Without templates, writing C++ functions that accept flexible inputs
is quite difficult because C++ is a statically-typed language. Usually one would
write one absolute function for doubles and another for integers whereas here we
don't have to.

### Notes on templates

To correctly register templates, the '[[cppally::register]]' tag must
always go above the function name.

```cpp
template <typename T>
[[cppally::register]] // <--- Here
T foo(T x){
  return x;
}
```

Explicit instantiation (from R) is unfortunately not possible and template
types must be deduced from supplied arguments.

```cpp
template <typename T>
[[cppally::register]]
T foo(){
	return T();
}
```

You may get a cryptic compiler error like this

```cpp
error: no matching function for call to 'foo()'
[]<typename T>() -> decltype(cpp_to_sexp(::foo())) {
```

along with an equally cryptic note

```cpp
note:   couldn't deduce template parameter 'T'
[]<typename T>() -> decltype(cpp_to_sexp(::foo())) {
```

This is because the parameter `T` cannot be automatically deduced from any
of the function inputs.
Even though these kinds of templates can be written with cppally,
they cannot be exported to R.

An obvious and somewhat ugly workaround is to include a prototype argument that
allows the template parameter to be deduced from.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["scalar_default"]])
```


```{r}
scalar_default(integer(1)) # Default is 0L
scalar_default(numeric(1)) # Default is 0.0
scalar_default(character(1)) # Default is ""
```


Exporting variadic templates are also not supported. The best alternative is
to use lists (`r_vec<r_sexp>`).

In the above example we used the `RScalar` concept which includes
all cppally scalar types (excluding `r_sexp`).
For a list of all cppally concepts, please see the **Annex**

## Coercion

To coerce from one scalar to another we can use `as<T>`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["double_to_int"]])
```

```{r}
double_to_int(pi)
double_to_int(NA_real_)
```

We can also coerce from one vector type to another

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["to_int_vec"]])
```

```{r}
to_int_vec(c(0, 1.5, NA))
```

Since `as<T>` is extremely flexible, we can also coerce
from a scalar to a vector or vice versa

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["coercions"]])
```

```{r}
coercions()
```

## Strings

cppally provides the useful string type `r_str`

We can create R strings easily

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_str_hello"]])
```

```{r, echo=FALSE}
r_str_hello()
```

To get a C or C++ string, use the members `c_str()` and `cpp_str()` respectively

C string via `c_str()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_str_hello_c_str"]])
```

```{r, echo=FALSE}
r_str_hello_c_str()
```

C++ string_view via `cpp_str()`

This can be converted into a std::string via its constructor

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["str_concatenate"]])
```

```{r}
str_concatenate("hello", "how are you?", sep = ", ")
```

## Symbols

Symbols have class `r_sym` and can be created directly from a string literal

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_sym_new"]])
```

```{r, echo=FALSE}
r_sym_new()
```

Or from a cppally string

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["r_sym_from_str"]])
```

```{r, echo=FALSE}
r_sym_from_str()
```

## Cached strings & symbols

cppally provides an efficient caching strategy for constructing cppally
strings/symbols from string literals

`cached_str<>`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["cached_str_demo"]])
```

```{r, echo=FALSE}
cached_str_demo()
```

This initialises the string once, caches it (to R's CHARSXP pool), and efficiently
re-uses the cached string for each subsequent call.

We can cache symbols in a similar way

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(single_exprs[["cached_sym_demo"]])
```

```{r, echo=FALSE}
cached_sym_demo()
```

## Lists

`r_sexp` is generally interpreted as an "element of a list" since lists are defined as `r_vec<r_sexp>`, a vector that holds generic `r_sexp` elements.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["new_list"]])
```

```{r}
new_list(0)
new_list(3)
```

The problem with a class like `r_sexp` is that it is by design
generic and therefore difficult to work with in C++.
To disambiguate the actual type we can use `visit_vector()` or `visit_sexp()`
via a C++ lambda.

**Example:** using `visit_vector()` to resize every vector to length n in-place

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["resize_all"]])
```

```{r}
# Resize to size 1
resize_all(list(1:5, letters), n = 1)
```

When we pass a non-vector to `visit_vector`, it aborts and
explains that the input must be a vector

```{r, error=TRUE}
resize_all(list(mean_fn = mean), 1)
```


**visit_sexp**

This allows us to visit more types than just vectors, including factors, symbols and (soon to be implemented) data frames. When an object's type can't be deduced into a distinct type, `r_sexp` is returned.

**Example:** Same example as above but with `visit_sexp()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["resize_all2"]])
```

```{r}
# Resize to size 1
resize_all2(list(1:5, letters), n = 1)
```

```{r, error=TRUE}
resize_all2(list(mean_fn = mean), n = 1)
```


## Factors

We can create a factor via `r_factors()`

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["new_factor"]])
```

```{r}
new_factor(letters)
```

In cppally, like R, factors are not vectors and therefore do not satisfy the RVector concept. To access the underlying integer codes vector, use the public `codes()` member function

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["factor_codes"]])
```

```{r}
letter_fct <- new_factor(letters)

letter_fct |>
    factor_codes()
```


## Attributes

Attributes can be manipulated via functions defined in the attr namespace.

**Example:** Adding names to a list

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(examples[["setnames"]])
```


```{r}
set.seed(42)
norm_samples <- lapply(1:5, \(x) rnorm(10, mean = x))
set_list_names(norm_samples, paste0("sample_", 1:5))
```


More useful attribute helpers

- `get_attrs()` - Returns a list of attributes (possibly `r_vec<r_sexp>(r_null)`)
- `set_attrs()` - Sets attributes to ones specified. Note: replaces any current attributes
- `clear_attrs()` - Removes all attributes
- `set_attr()` - Set a single attribute
- `get_attr()` - Get a single attribute
- `inherits1()` - Does object inherit class?
- `inherits_any()` - Does object inherit at least one of the specified classes?
- `inherits_all()` - Does object inherit all of the specified classes?
- `modify_attrs()` - Modifies current attributes but doesn't remove any existing ones

## Sugar functions

cppally also offers many useful and high-performance common functions in cppally/sugar

**Example:** `n_unique()` - fast calculation of number of unique values.

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(benchmark_examples[["cpp_n_unique"]])
```

```{r}
library(bench)
x <- sample(1:100, 10^5, replace = TRUE)
mark(
  base_n_unique = length(unique(x)),
  cppally_n_unique = cpp_n_unique(x)
)
```


More useful sugar functions

- `unique()` - Like R's `unique()` but with a `sort` argument to return sorted unique values

- `identical()` - A very fast identical function that works for scalars and vectors. Use this for exact equality of any scalar or vector.

- `match()` - Like R's match, but also faster

- `sequences()` - Like `sequence()` but it returns a list of sequences and also works with doubles.

- `order()` - Like base R's order but it internally uses a hybrid approach of ska sort, count sorting, quick sort, etc.

- `make_groups()` - An advanced function that returns a struct containing group IDs and number of groups (i.e number of unique group IDs). The `groups` struct contains the following members:
	- r_vec<r_int> ids - The cached group IDs
	- int n_groups - Number of unique groups
	- bool ordered - Do the group IDs specify a sorting order, or are they by order-of-first-appearance?
	- bool sorted - Are the group IDs sorted? (This can also be true for order-of-first-appearance IDs)
	- r_vec<r_int> start() - Returns an r_vec<r_int>(n_groups) vector of start locations of each unique group, signifying the location in the data at which each group initially appeared
	- r_vec<r_int> counts() - Returns an r_vec<r_int>(n_groups) vector of frequency counts of each unique group
	- r_vec<r_int> order() - Returns an r_vec<r_int>(ids.length()) order vector. This is a 0-indexed permutation vector that can be used to return sorted group IDs

- `recycle()` - Recycles supplied vectors to common length

- `r_vec<T>::subset()` - Fast subsetting of vectors

**Scalar math functions**

There is a rich suite of math functions. Some examples include `min()`, `max()`, `round()`, `log()`, `floor()`, `ceiling()` and more.

**Stats sugar functions**

Some statistical summary functions that are all very highly optimised for speed

- `sum()` - Sum of values
- `range()` - Min and max range of values
- `abs()` - Computes absolute values (there is also a scalar version)
- `var()` and `sd()` - Variance and standard deviation
- `gcd()` - Greatest common divisor
- `lcm()` - Lowest common multiple

## Annex

### Symbols in R-registered templates

`r_sym` is unsupported in templates when it's part of a template argument but is supported when the argument is explicitly an `r_sym`.

```cpp
[[cppally::register]]
r_str symbol_to_string(r_sym x){
	return as<r_str>(x);
}
```

```r
hello_world_symbol <- as.symbol("hello world!")
hello_world_symbol
`hello world!`
symbol_to_string(hello_world_symbol)
[1] "hello world!"
```

### All core cppally concepts

- RIntegerType - Includes `r_lgl`, `r_int`, `r_int64`

- RMathType - Includes `r_lgl`, `r_int`, `r_int64` and `r_dbl`

- RStringType - Includes `r_str` and `r_str_view`

- RScalar - Includes all cppally specific scalar types

- RVal - Includes anything a cppally vector (`r_vec<>`) can contain: RScalar +`r_sexp`

- RVector - Includes `r_vec<T>` where `T` is an RVal
- RTimeType - Includes `r_date` and `r_psxct`
- RNumericType - Numeric types, including RMathType and RTimeType
- RSortableType - Includes RNumericType and RStringType (strings can also be sorted)
- RAtomicVector - A vector that contains RScalar elements
- CppallyType - Any R type defined by R, including RVal, RVector, RFactor, RDataFrame, RSymbol
- CppType - Anything that is not an CppallyType
- CastableToRScalar - Anything that can be constructed or cast into an RScalar (which also includes RScalar)
- CastableToRVal (**questioning**) - Anything that can be constructed or cast into an RVal. This is more complicated as it includes vectors, factors and data frames which can be cast to `r_sexp`

Other useful type traits

- `unwrap_t` - Returns the underlying unwrapped type
- `as_r_scalar_t` - Returns the equivalent RScalar type
- `as_r_val_t` - Returns the equivalent RVal type
- `common_r_t` - Returns the common RVal type between 2 types. Generally this is a hierarchy where the common type is the type that both values can be coerced to without complete loss of information


### Accessing the underlying types and values

While it is generally recommended not to access the underlying objects, you can do so with `unwrap()` which returns the underlying C/C++ value. For example, `unwrap(r_int(5))` will return an `int` of value `5`.

To access the underlying type, use `unwrap_t<>` which always aligns with `unwrap()`

The main reason for wanting to access underlying values would likely be optimisation and so `unwrap()` and `unwrap_t` allow this to be done consistently.

**Example:** Summing a double vector using `r_vec<T>::data()` member

```{r, echo=FALSE, results = 'asis'}
as_cpp_chunk(benchmark_examples[["primitive_sum"]])
```

```{r}
x <- rnorm(10^5)
primitive_sum(x)
```
