---
title: "Working with lists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with lists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

```{r, include = FALSE}
required <- c("bench", "brio", "callr", "cli", "decor",
              "desc", "glue", "purrr", "readr", "stringr",
              "utils", "vctrs", "withr")
if (!all(vapply(required, requireNamespace, logical(1), quietly = TRUE))) {
  knitr::knit_exit()
}
```

Lists are R's generic vector class, capable of holding any object. 
Every other vector holds elements of a single known type, 
but a list can hold anything — including other
lists. This vignette explains how cppally represents lists, why that
representation falls naturally out of the type system, and how to bridge the
gap between R's dynamically typed lists and C++'s statically typed world.

```{r}
library(cppally)
```

```{r, include=FALSE}
# Helpers to compile all examples in debug mode
cpp_source <- function(..., code, debug = TRUE, env = parent.frame()){
  preamble <- c("#include <cppally.hpp>", "using namespace cppally;")
  code <- paste(c(preamble, code), collapse = "\n")
  cppally::cpp_source(debug = debug, env = env, code = code, ...)
}

# Helpers to source and display C++/R code
chunk_impl <- function(x, language){
  paste0("```", language, "\n", x, "\n```\n")
}
as_code_chunk <- function(x, language){
  cat(chunk_impl(x, language))
}
as_cpp_chunk <- function(x){
  as_code_chunk(x, "cpp")
}
```

```{r, include=FALSE}
# Compile necessary examples in one-go
# as it's faster when building the vignette

examples <- c(
  new_list = '
[[cppally::register]]
r_vec<r_sexp> new_list(int n){
  return r_vec<r_sexp>(n);
}
',
  list_lengths = '
[[cppally::register]]
r_vec<r_int> list_lengths(r_vec<r_sexp> x){
  r_size_t n = x.length();
  r_vec<r_int> out(n);
  for (r_size_t i = 0; i < n; ++i){
  
    r_sexp list_elem = x.get(i);
    
    r_int elem_length = r_sexp_visit(list_elem, []<RVector V>(const V& v){
      return as<r_int>(v.length());
    });
    
    out.set(i, elem_length);

  }
  return out;
}
',
  times_two = '
[[cppally::register]]
r_sexp times_two(r_sexp x){
  r_sexp_mutate(x, [](r_vec<r_dbl>& v){
    r_size_t n = v.length();
    for (r_size_t i = 0; i < n; ++i){
      v.set(i, v.get(i) * 2);
    }
  });
  return x;
}
',
  list_times_two = '
[[cppally::register]]
r_vec<r_sexp> list_times_two(r_vec<r_sexp> x){
  r_size_t n = x.length();
  for (r_size_t i = 0; i < n; ++i){

    r_sexp elem = x.get(i);

    r_sexp_mutate(elem, [](r_vec<r_dbl>& v){
      r_size_t m = v.length();
      for (r_size_t j = 0; j < m; ++j){
        v.set(j, v.get(j) * 2);
      }
    });

    x.set(i, elem);

  }
  return x;
}
',
  my_rev = '
template <RVector T>
T my_rev(const T& x){
  r_size_t n = x.length();
  T out(n);
  for (r_size_t i = 0; i < n; ++i){
    out.set(i, x.get(n - i - 1));
  }
  return out;
}

r_sexp my_rev(const r_sexp& x){
  return r_sexp_visit(x, []<RVector V>(const V& v){
    return static_cast<r_sexp>(my_rev(v));
  });
}
',
  rev_registered = '
[[cppally::register]]
r_sexp cpp_rev(r_sexp x){
  return my_rev(x);
}

[[cppally::register]]
r_vec<r_sexp> rev_list(r_vec<r_sexp> x){
  r_size_t n = x.length();
  r_vec<r_sexp> out(n);
  for (r_size_t i = 0; i < n; ++i){
    out.set(i, my_rev(x.get(i)));
  }
  return out;
}
'
)

cpp_source(code = paste(examples, collapse = "\n"))
```

## Everything is a SEXP

In R's C API, every R object — an integer vector, a list, a function, an
environment, even `NULL` — is represented by a single C type: `SEXP`. A `SEXP`
is a pointer to R's internal object representation, and the actual type of the
object it points to is stored inside the object itself, discovered at runtime
with `TYPEOF()`.

In other words, a `SEXP` is dynamically typed, just like R. This is what makes
it so flexible: any C function that accepts a `SEXP` can, in principle, accept
any R object. The flip side is that a raw `SEXP` offers no compile-time
guarantees about what it holds, and any `SEXP` you allocate must be manually
protected from R's garbage collector.

## cppally's `r_sexp`

`r_sexp` is cppally's answer to the raw `SEXP`: a thin wrapper that keeps all
of the flexibility while removing the manual protection burden. When you
construct an `r_sexp` from a `SEXP`, the object is automatically inserted into
cppally's protection pool and released again when the last `r_sexp` pointing
to it is destroyed. Copies are reference-counted and practically free. See
`vignette("protection")` for the details and benchmarks of this design.

Crucially, `r_sexp` inherits the dynamic nature of `SEXP`: an `r_sexp` can
hold any R object. It also converts implicitly back to `SEXP`, so it can be
passed anywhere a `SEXP` is expected.

## Fixed scalar and vector types

Most of cppally lives at the opposite end of the spectrum. The scalar types
(`r_lgl`, `r_int`, `r_dbl`, `r_str`, and friends) and the vector template
`r_vec<T>` are statically typed: the element type is fixed at compile time.
This is what you want most of the time — a signature like

```cpp
r_dbl cpp_sum(r_vec<r_dbl> x)
```

documents exactly what the function accepts, lets the compiler generate fast
type-specific code, and turns type mistakes into compile-time errors rather
than runtime surprises.

## Lists are generic vectors

So where do lists fit in? At the C level, a list is a *generic vector*
(`VECSXP`): a vector whose elements are themselves `SEXP`s. Each element can
be any R object, so no fixed C++ element type can describe a list's contents —
except the one type designed to hold any R object: `r_sexp`.

This is exactly how cppally represents lists. There is no separate list class;
a list is simply

```cpp
r_vec<r_sexp>
```

a vector whose elements are dynamically typed handles. All the usual `r_vec`
machinery applies, including initialisation with `r_vec<T>(n)`:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["new_list"]])
```

```{r}
new_list(3)
```

You can also use an alias, such as `list`, for stylistic preference.

```cpp
using list = r_vec<r_sexp>;

[[cppally::register]]
list new_list(int n){
  return list(n);
}
```

Just note that a `list` alias will clash with `std::list` if you also have
`using namespace std;` in the same translation unit.

## Bridging the gap with `r_sexp_visit()`

Building lists is the easy part. The moment you pull an element back *out* of
a list, you are handed an `r_sexp` — a dynamically typed handle — and the
statically typed world has no idea what to do with it. You cannot sum it,
compare its elements or check them for `NA`, because none of those operations
make sense until the concrete type is known.

The concrete type *is* known — but only at runtime, inside the object. What we
need is a mechanism that inspects the runtime type and hands us back the
matching statically typed wrapper. That mechanism is `r_sexp_visit()`:

```cpp
r_sexp_visit(x, []<typename T>(const T& v){
  // `v` is a concrete typed wrapper of type `T`, e.g. r_vec<r_int> or r_vec<r_str>
});
```

You pass an `r_sexp` and a generic function (typically a template lambda).
`r_sexp_visit()` checks the runtime type of the underlying object and invokes
your function with the corresponding cppally wrapper — `r_vec<r_lgl>`,
`r_vec<r_int>`, `r_vec<r_dbl>`, `r_vec<r_str>`, and so on, including
`r_vec<r_sexp>` itself for nested lists, and `r_factors` or `r_df` for
factors and data frames.

The concept on the lambda's template parameter constrains which types 
your visitor accepts. 

```cpp
r_sexp_visit(x, []<RVector T>(const T& v){
  // Try visit `x` as an RVector
});
```

If the object's runtime type is not one the
visitor accepts, `r_sexp_visit()` aborts with an error at runtime — the
dynamic world's version of a compile-time type error.

Here is the classic first example, an implementation of R's `lengths()`:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["list_lengths"]])
```

```{r}
list_lengths(list(1:10, letters, NULL, list("a", 1)))
```

For each element we visit its concrete type and ask for its length. The
visitor is instantiated once per possible wrapper type at compile time, so
inside the lambda you get full static typing — the dynamic dispatch happens
exactly once, at the visit itself.

Since we constrained on `RVector`, visited types must satisfy that and can't be 
for example, a data frame.

```{r,error=TRUE}
list_lengths(list(data.frame(x = 1:3)))
```

`r_sexp_visit()` also has a mutating sibling, `r_sexp_mutate()`, which the
next section covers.

## Modifying in place with `r_sexp_mutate()`

`r_sexp_mutate()` hands your visitor a
*mutable* typed wrapper, and any modification the visitor makes lands back
in the `r_sexp` you passed in. Note the signature: it takes its argument by
`r_sexp&`, not `const r_sexp&`, because under the hood it moves the object
into the typed wrapper, runs your visitor on it, and then writes the
wrapper's (possibly replaced) `SEXP` back out.

A visitor doesn't have to be a template, either. A plain lambda taking
exactly one wrapper type is a perfectly good visitor — it simply constrains
the accepted types to that single type:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["times_two"]])
```

```{r}
times_two(c(1.5, 2.5, 3.5))
```

Any other type is rejected, and the error message tells us exactly what this
visitor is prepared to accept:

```{r, error=TRUE}
times_two(letters)
```

Applied to lists, `r_sexp_mutate()` lets us modify each element of a list in
turn:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["list_times_two"]])
```

```{r}
list_times_two(list(c(1.5, 2.5), c(30, 40)))
```

Note the pull-out / mutate / write-back rhythm: we `get()` the element,
mutate it, then `set()` it back into the list. The write-back matters —
`r_sexp_mutate()` is allowed to *replace* the element's underlying `SEXP`
rather than modify it (for instance when the object is shared and must be
duplicated first to preserve R's copy-on-modify semantics), and the final
`set()` is what makes the list point at the result.

## Writing methods that work on `r_sexp`

cppally ships a family of free functions that work on any cppally type —
`length()`, `rep_len()`, `rep()`, `subset()`, `order()`, `n_unique()` and
friends. Each follows the same two-part pattern: a set of typed overloads
that do the real work, plus an `r_sexp` overload that recovers the concrete
type and forwards. Here is the essence of `length()`, simplified:

```cpp
// Typed: one template covers every vector
template <RVector T>
r_size_t length(const T& x){
  return x.length();
}

// Dynamic: recover the type, then reuse the typed overload
r_size_t length(const r_sexp& x){
  return r_sexp_visit(x, []<RVector V>(const V& v){
    return length(v);
  });
}
```

For a function that *produces* a vector, like `rep_len()`, there is one
extra wrinkle: the
visitor's arms must all agree on a single return type, and when the input
could be any vector type, the output could be too. The type that can hold
any result is, once again, `r_sexp`:

```cpp
template <RVector T>
T rep_len(const T& x, r_size_t n){
  return x.rep_len(n);
}

r_sexp rep_len(const r_sexp& x, r_size_t n){
  return r_sexp_visit(x, [&]<RVector V>(const V& v){
    return static_cast<r_sexp>(rep_len(v, n));
  });
}
```

There is a pleasing symmetry to it: typed in, typed out; dynamic in, dynamic
out.

Writing your own method in this style is no different. Here is a reverse
function — a typed template that does the work, plus the `r_sexp` overload
that bridges:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["my_rev"]])
```

Neither of these is exposed to R directly — they are building blocks. We
register two thin wrappers instead:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["rev_registered"]])
```

```{r}
cpp_rev(c(1.5, 2.5, 3.5))
cpp_rev(list(1:3, letters[1:3]))
rev_list(list(1:3, letters[1:3]))
```

Two things worth noticing. `cpp_rev()` on a list reverses the *order of the
elements* — the visit dispatches the list to the `r_vec<r_sexp>` arm of the
template, where elements are shuffled around as opaque handles, exactly like
R's `rev()`. And `rev_list()` shows the payoff of the `r_sexp` overload for
list processing: each element comes out of the list as an `r_sexp`, and
`my_rev(x.get(i))` just works on every one of them, whatever they hold —
the dispatch is tucked away in the overload, and the calling code stays one
line per element.
