---
title: "Random number generation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random number generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

```{r, include = FALSE}
required <- c("bench", "brio", "callr", "cli", "decor",
              "desc", "glue", "utils", "withr")
if (!all(vapply(required, requireNamespace, logical(1), quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
  knitr::knit_exit()
}
```

In many analyses, reproducibility is just as important as the quality of 
the analytical methods used. The ability to reproduce the same set of numbers 
is a core mechanism of R. 

`random_stream` exists to allow that reproducibility from R, but entirely in C++.

```{r}
library(cppally)
```

```cpp
#include <cppally.hpp>
using namespace cppally;
```

```{r, include=FALSE}
# Helpers to compile all examples in debug mode
cpp_source <- function(..., code, debug = TRUE, env = parent.frame()){
  preamble <- c(
    "#include <cppally_light.hpp>", 
    "#include <cppally/random/random_stream.h>", 
    "#include <cppally/sugar/r_stats.h>", 
    "using namespace cppally;"
    )
  code <- paste(c(preamble, code), collapse = "\n")
  cppally::cpp_source(debug = debug, env = env, code = code, ...)
}

# Helpers to source and display C++/R code
chunk_impl <- function(x, language){
  paste0("```", language, "\n", x, "\n```\n")
}
as_code_chunk <- function(x, language){
  cat(chunk_impl(x, language))
}
as_cpp_chunk <- function(x){
  as_code_chunk(x, "cpp")
}
```

```{r, include=FALSE}
# Compile necessary examples in one-go
# as it's faster when building the vignette

examples <- c(
  one_unif = '
[[cppally::register]]
r_dbl one_unif(){
  return r_dbl(random_stream().unif());
}
',
  unif_deviates = '
[[cppally::register]]
r_vector<r_dbl> unif_deviates(int n){

  random_stream r; // <-- Generated once here

  r_vector<r_dbl> out(n);

  for (r_size_t i = 0; i < n; ++i){
    out.set(i, r.unif());
  }
  return out;
}
',
  unif_seeded = '
[[cppally::register]]
r_vector<r_dbl> unif_seeded(int n, uint64_t seed){

  random_stream r(seed);

  r_vector<r_dbl> out(n);

  for (r_size_t i = 0; i < n; ++i){
    out.set(i, r.unif());
  }
  return out;
}
',
  normal_deviates = '
#include <random>

[[cppally::register]]
r_vector<r_dbl> normal_deviates(int n){

  random_stream r;
  r_vector<r_dbl> out(n);

  std::normal_distribution<> dist;

  for (r_size_t i = 0; i < n; ++i){
    out.set(i, dist(r));
  }
  return out;
}
',
  r_normal_deviates = '
[[cppally::register]]
r_vector<r_dbl> r_normal_deviates(int n){

  r_vector<r_dbl> out(n);

  draw_from_r([&]{
    for (r_size_t i = 0; i < n; ++i){
      double v = norm_rand();
      out.set(i, v);
    }
  });

  return out;
}
',
  random_index = '
[[cppally::register]]
int random_index(int a, int b){
  return random_stream().index(a, b);
}
'
)

cpp_source(code = paste(examples, collapse = "\n"))
```

```{r, include=FALSE}
# Benchmarked separately, in an optimised build

benchmark_examples <- c(
  cpp_sample_int_with_replacement = '
[[cppally::register]]
r_vector<r_int> cpp_sample_int_with_replacement(int n, int size){

  random_stream r;
  r_vector<r_int> out(size);

  for (r_size_t i = 0; i < size; ++i){
    auto idx = r.index(0, n - 1);
    out.set(i, idx);
  }
  return out;
}
',
  boot_mean = '
[[cppally::register]]
r_vector<r_dbl> boot_mean(r_vector<r_dbl> x, int n_boot){

  r_size_t n = x.length();

  random_stream r;
  r_vector<r_dbl> out(n_boot);

  for (int b = 0; b < n_boot; ++b){
    
    r_dbl total(0);
    
    // calculate mean
    for (r_size_t i = 0; i < n; ++i){
      total += x.get(r.index(r_size_t(0), n - 1));
    }
    out.set(b, total / n);
    
  }
  return out;
}
',
resample = '
[[cppally::register]]
r_vector<r_dbl> resample(r_vector<r_dbl> x){

    r_size_t n = x.length();
    r_vector<r_dbl> out(n);
    
    random_stream r;
    
    for (r_size_t i = 0; i < n; ++i){
        out.set(i, x.get(r.index(r_size_t(0), n - 1)));
    }
    
    return out;
}
',
apply_boot = '
template <RScalar T, typename FUN>
requires RScalar<std::invoke_result_t<FUN&, const r_vector<T>&>> // Accept only fns that return scalars
auto apply_boot(const r_vector<T>& data, int n_boot, FUN fn){

  r_size_t n = data.length();

  // Function return type
  using stat_t = std::invoke_result_t<FUN&, const r_vector<T>&>;

  random_stream r;

  r_vector<stat_t> out(n_boot);
  
  // Reused across every replicate
  r_vector<T> samples(n);

  for (int b = 0; b < n_boot; ++b){

    // Sample x using a random index [0, n - 1]
    // Set the samples directly into our existing samples vector
    for (r_size_t i = 0; i < n; ++i){
        
        auto idx = r.index(r_size_t(0), n - 1);
        
        // data.view() is safe here because we are immediately protecting the result from gc
        // by placing it into samples, which is itself protected
        
        samples.set(i, data.view(idx));
    }

    // Apply function
    out.set(b, fn(samples));

  }

  return out;
}
',
boot_mean2 = '
[[cppally::register]]
r_vector<r_dbl> boot_mean2(r_vector<r_dbl> x, int n_boot){
    return apply_boot(x, n_boot, [](const auto& vec){
        return mean(vec);
    });
}
',


boot_stats = '

// Bootstrapped sum

[[cppally::register]]
r_vector<r_dbl> boot_sum(r_vector<r_dbl> x, int n_boot){
    return apply_boot(x, n_boot, [](const auto& vec){
    
      return as<r_dbl>(sum(vec));
        
    });
}

// Bootstrapped count of positive values

[[cppally::register]]
r_vector<r_int> boot_n_positive(r_vector<r_dbl> x, int n_boot){
    return apply_boot(x, n_boot, [](const auto& vec){
    
      return vec.reduce([](auto acc, auto curr){
        return as<r_int>(acc + (curr > 0));
      }, 
      /*init = */ r_int(0));
        
    });
}
'
)

cpp_source(code = paste(benchmark_examples, collapse = "\n"), debug = FALSE)
```

## Creating a stream

The default constructor seeds a new stream from R's *current* seed state:

```cpp
random_stream r; // Initialised with a new seed using R's current seed state
```

The consequence is the one that matters for users: a function that builds a
stream is reproducible from R.

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["one_unif"]])
```

```{r}
set.seed(1)
one_unif()

set.seed(1)
one_unif() # Reproduces the same number under the same R seed
```

Without setting a seed, we get random numbers as expected.

```{r}
replicate(10, one_unif())
```

Because constructing a stream *consumes* R draws, two fresh streams under one
`set.seed()` are different - exactly as you would want, and exactly like
calling `runif()` twice:

```{r}
set.seed(1)
c(one_unif(), one_unif())
```

## Reusing one stream

When doing repeated sampling, it's more efficient to generate the
`random_stream` once and draw from it many times, rather than constructing a
new one per draw.

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["unif_deviates"]])
```

```{r}
set.seed(1)
unif_deviates(10)
```

`unif()` returns a `double` in `[0, 1)` by default, but can return any range 
in `[a, b)`.

## Explicit seeds

A stream can also be built from a seed you choose, in which case R's state is
irrelevant and never touched:

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["unif_seeded"]])
```

```{r}
set.seed(1)
unif_seeded(5, seed = 42)

set.seed(99)
unif_seeded(5, seed = 42) # Same numbers, different R seed
```

Every stream can return the seed it was created with, so a run seeded from R can
still be replayed later:

```cpp
random_stream r;
uint64_t s = r.seed();   // Log this
// ... later ...
random_stream replay(s); // Identical sequence
```

## `index()`

`random_stream::index()` samples a random integer between the specified
bounds (**inclusive** at both ends).

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["random_index"]])
```

```{r}
set.seed(1)
random_index(0, 1000)
```

Unlike the `<random>` distributions discussed below, `index()` is
platform-independent. It uses Lemire's nearly divisionless method over pure
integer arithmetic, so the same seed gives the same integers everywhere.

### Example: efficient sample with replacement

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(benchmark_examples[["cpp_sample_int_with_replacement"]])
```

Benchmark comparison against `sample.int(..., replace = TRUE)`:

```{r}
library(bench)
mark(
  cppally_sample_int_with_replacement = cpp_sample_int_with_replacement(10^6, 10^5),
  base_sample_int_with_replacement = sample.int(10^6, 10^5, replace = TRUE),
  check = FALSE
)
```

In this simple benchmark we achieve a large speed improvement over base R.
The whole loop runs without re-entering R's RNG once - only the
`random_stream r;` line does that.

## Using `<random>` distributions

`random_stream` satisfies `std::uniform_random_bit_generator`, so it can be
handed straight to any C++ distribution or algorithm - `std::shuffle`,
`std::normal_distribution` and so on.

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["normal_deviates"]])
```

```{r}
set.seed(1)
normal_deviates(10)
```

Two caveats are worth knowing before you rely on this.

The C++ standard specifies the *interface* of the `<random>` distributions but
not their algorithms, so results are not guaranteed to be identical across
platforms, compilers, or standard library versions. The same seed reproduces
the same numbers on one machine, but not necessarily on somebody else's.
Relatedly, *Writing R Extensions* asks compiled code not to depend on the
C++11 random number library - which is why `random_stream`'s own methods
(`unif()`, `index()`) are implemented directly rather than delegating to
`<random>`.

If cross-platform reproducibility matters, prefer `unif()`/`index()`, or R's
own distribution functions via `draw_from_r()`.

## Calling R C API RNG functions

⚠️ To call R's C API RNG functions - `unif_rand()`, `norm_rand()`,
`R_unif_index()`, or anything from `Rmath.h` - wrap them in
`draw_from_r()`.

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(examples[["r_normal_deviates"]])
```

```{r}
set.seed(1)
r_normal_deviates(5)

set.seed(1)
rnorm(5) # Identical - these are R's own draws
```

`draw_from_r()` loads R's RNG state before running your callable and writes
it back after, i.e. it is an RAII wrapper around `GetRNGstate()` and
`PutRNGstate()`. Without it, draws never reach `.Random.seed`, so R hands the
same values out again on its next call.

It returns whatever the callable returns, so it wraps expressions as well as
statements:

```cpp
uint64_t seed = draw_from_r([]{ return internal::draw_seed(); });
```

Wrap the *whole loop*, as above, rather than each individual draw - the
state round trip is the expensive part.

## Bootstrapping

Given that we can generate random indices very efficiently, it stands 
to reason that we can also generate bootstrap samples efficiently.

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(benchmark_examples[["boot_mean"]])
```

`boot_mean` resamples `x` (with replacement), generating `n_boot` mean estimates. 
We can then use `quantile` to create a percentile confidence interval.

```{r}
set.seed(42)
x <- rnorm(1000)

set.seed(1)
replicates <- boot_mean(x, 2000)

quantile(replicates, c(0.025, 0.975)) # Percentile confidence interval
```

### Comparison to R - benchmark

We could compare `boot_mean` to R's equivalent `mean` + `sample`, but that 
wouldn't be a fair comparison, given that `boot_mean` is a hand-tuned C++ 
function which doesn't generate intermediate vectors. 

Instead let's compare apples to apples and write our own C++ sample-with-replacement.

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(benchmark_examples[["resample"]])
```

Now let's test it against R. 

```{r}
boot_mean_r <- function(x, n_boot){
  vapply(
    seq_len(n_boot),
    \(i) mean(sample(x, replace = TRUE)), 
    0
  )
}

boot_mean_cppally <- function(x, n_boot){
  vapply(
    seq_len(n_boot),
    \(i) mean(resample(x)),
    0
  )
}

mark(
  cppally = boot_mean_cppally(x, 2000),
  base_r  = boot_mean_r(x, 2000),
  check = FALSE
)
```

The hand-tuned bootstrap mean naturally is faster.

```{r}
mark(
  boot_mean_cppally(x, 2000),
  boot_mean(x, 2000),
  check = FALSE
)
```

### A flexible and fast bootstrapper

The gap between `boot_mean_cppally` and `boot_mean` is worth pulling apart. Three
costs are paid on every one of those 2,000 iterations: `resample()` builds a fresh
`random_stream`, so R's `.Random.seed` is loaded and written back each time -
exactly what *Reusing one stream* set out to avoid; a new `n`-length vector is
allocated only to have a single number taken from it; and each replicate makes a
round trip out to R and back.

`boot_mean` pays none of them per replicate. It constructs one stream, allocates
nothing inside the loop, and never leaves C++.

All three go away together if we hold one stream for the whole run, re-use one
intermediate vector, and keep the replicate loop in C++. If we can manage that
without fixing the statistic in advance, we get something generic that still
performs like the hand-tuned version.

**Bootstrap functional:**

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(benchmark_examples[["apply_boot"]])
```

Most of the complexity sits in the signature. The three arguments are:

- `data` - the vector being bootstrapped
- `n_boot` - the number of replicates
- `fn` - a C++ lambda returning the test-statistic of your choice

The `requires` clause rejects any `fn` that does not return a scalar, and
`stat_t` picks up whatever type it *does* return.

Using cppally's `cppally::mean()`, we can easily re-write a new mean bootstrap 
function. 

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(benchmark_examples[["boot_mean2"]])
```

Comparing our new bootstrap mean function against the hand-tuned version.

```{r}
mark(
  boot_mean(x, 2000),
  boot_mean2(x, 2000),
  check = FALSE
)
```

Almost as fast as the hand-tuned version, which is a nice result given that 
`boot_mean2` was written using a functional.

One benefit of a generic bootstrap functional like `apply_boot` 
is the ability to quickly write new bootstrap methods. 

```{r, echo=FALSE, comment="", results='asis'}
as_cpp_chunk(benchmark_examples[["boot_stats"]])
```

Note that these two return different types. `boot_sum` produces doubles, whereas
`boot_n_positive` counts, and so produces integers. Neither had to declare that -
`stat_t` follows whatever the lambda returns, and the output vector is built from
it.

```{r}
set.seed(42)
boot_sum(x, 5)
boot_n_positive(x, 5)
```

## Engine

`random_stream` drives xoshiro256++ (Blackman & Vigna, public domain) via
Xoshiro-cpp v1.0 by Ryo Suzuki (MIT), bundled in `inst/include/` and reduced
to xoshiro256++ and its SplitMix64 seeder.
