---
title: "Vector Names Hashing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vector Names Hashing}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To speed up named-indexing, cppally implements hashing of 
vector names via a lazy cache-on-lookup approach. To better understand how 
this works, we will show what name-indexing in R looks like, how to improve it 
with hashing, as well as the benefits and limitations of cppally's inherent 
lazy-hashing method.

Let's first load cppally

```{r setup}
library(cppally)
```

Indexing into a vector by name is a useful idiom

```{r}
# Name-value list
x <- list(
  a = 10, 
  b = 20, 
  c = 30
)

x[["c"]] # Index-by-name
x[[3]] # Index-by-location 
```

It signifies to those reading the code the exact key we want. Compared
to indexing by location, it is resilient to keys changing location.

```{r}
# If we insert an element between 2 and 3, x[[3]] changes
x <- c(
  x[1:2], 
  list(d = 40), # New element
  x[3]
)

x[[3]] # Now 40
x[["c"]] # Still the same value 
```

## Performance of indexing by location versus name

For this we need a large name-value list

```{r}
set.seed(42)
large <- as.list(sample.int(10^5))
names(large) <- paste0("name_", seq_along(large))
```

Lookup by location is O(1) and very fast in R, regardless of the index

```{r}
library(bench)
mark(large[[1]])
mark(large[[length(large)]])
```

Since R scans the names each time, lookup by name is O(n) and can be
slow if names happen to be near the end of the vector

```{r}
mark(
    by_name = large[["name_1"]], 
    by_index = large[[1]]
)

mark(
    by_name = large[["name_100000"]], 
    by_index = large[[100000]]
)
```

If we created a hash table of names-values, we could speedup repeated
lookups

```{r}
names_hashtab <- hashtab(size = length(large))
for (i in seq_along(names(large))){
 nm <- names(large)[[i]]
 sethash(names_hashtab, nm, large[[i]])
}

# Confirm it worked
identical(gethash(names_hashtab, "name_10"), large[["name_10"]])
```

Let's now compare all three methods: lookup by location, name and hashed
name

```{r}
mark(
    by_name = large[["name_100000"]], 
    by_index = large[[100000]],
    by_hashed_name = gethash(names_hashtab, "name_100000")
)
```

That worked! Extracting the value associated with the last name using
the hash table was almost as fast as looking it up by location.

cppally uses this same approach internally, 
attaching a lazy hash map cache to each `r_vector`.

Due to R/C++ limitations, the cache cannot survive the R/C++ boundary as
we cannot easily attach the cache to the returned `SEXP`. Therefore
instead of a cache-on-1st-lookup approach, which would be most
efficient, we implement a cache-on-2nd-lookup approach. This brings
certain limitations when writing R functions that use cppally code.

To understand the limitation, it is first crucial to understand how
cppally registers C++ functions to R. R can only handle `SEXP` objects, a
generic type that represents any R object.

cppally automatically handles all conversions between cppally classes
like `r_vector` and R's `SEXP`. To illustrate this, suppose we have the
identity function `foo()` that accepts an integer vector `x` and returns
the same vector unchanged.

``` cpp
[[cppally::register]]
r_vector<r_int> foo(r_vector<r_int> x){
  return x;
}
```

The function is then registered to R (via `cppally::load_all()`) and the
following R and C++ code is automatically generated.

### Generated R code

``` r
foo
function(x) {
  .Call(`_cppally_foo`, x)
}
```

### Generated C++ code

``` cpp
r_vector<r_int> foo(r_vector<r_int> x);
extern "C" SEXP _cppally_foo(SEXP x) {
  BEGIN_CPPALLY
  return cpp_to_sexp(::foo(as<r_vector<r_int>>(x)));
  END_CPPALLY
}
```

Looking at the generated R function `foo()`, we see `.Call()` is called
and is passed the input `x`. `.Call()` is the R interface that allows us
to pass R objects to C++.

Looking at the generated C++ function - `BEGIN_CPPALLY/END_CPPALLY` are
helpers that run the main code in try/catch blocks to catch any C++
errors and promote them to R errors. In theory, all R errors thrown by
cppally are done via `R_UnwindProtect` which safely runs all C++
destructors on error.

Looking at the innermost function, we can see the expression
`as<r_vector<r_int>>(x)` - this converts the input `SEXP` from R into a
fresh cppally integer vector. The C++ function `foo()` is then called
with that vector, which is then passed to `cpp_to_sexp()` - a function
that converts C/C++ objects back to `SEXP`.

The conversions look like this:

```
SEXP              (R)
  ↓ 
r_vector<r_int>   (C++)
  ↓
r_vector<r_int>   (C++, via foo())
  ↓
SEXP              (R)
```

This round-trip from `SEXP` to `r_vector<r_int>` and back to `SEXP`
happens every single time `foo()` is called (which happens via
`.Call()`). Because `r_vector` is the class that builds and stores the
cache, that cache is lost when converting to `SEXP` and therefore, the
hashing approach becomes difficult.

To mitigate the negative performance impacts of this limitation, we
implement a hash-on-2nd-lookup approach. This means the first lookup is
a linear scan, the second lookup triggers the hash map build before
doing a hash lookup. All lookups from the 2nd onwards use the hash map.

Practically speaking this shouldn't negatively impact performance when
doing lookups from R as one would have to write a function that performs
exactly two lookups on each call to experience the highest cost penalty,
something that while not impossible, is likely uncommon.

In C++, the cache persists across calls since there is no `SEXP`
round-trip. The first lookup is a linear scan, the second builds the
hash map, and all subsequent lookups are O(1).

``` cpp
r_int a = x.get("a"); // linear scan 
r_int b = x.get("b"); // hash map built + O(1) lookup 
r_int c = x.get("c"); // O(1) lookup
```

Let's illustrate this by visualising the per call cost as the number of
lookups increases, comparing R and cppally.

```{r}
cpp_source(code = '
#include <cppally.hpp>
using namespace cppally;

[[cppally::register]]
r_vec<r_sexp> do_lookup(r_vec<r_sexp> x, r_str name, int n_iterations){
  r_vec<r_sexp> out(n_iterations);
  for (int i = 0; i < n_iterations; ++i){
    out.set(i, x.get(name));
  }
  return out;
}
')

```

R equivalent using `[[`

```{r}
r_do_lookup <- function(x, name, n_iterations){
  out <- vector("list", n_iterations)
  for (i in seq_along(out)){
    out[[i]] <- x[[name]]
  }
  out
}
```

## The cost of first lookup

Here we are simply comparing the cost of doing a 1st-lookup, in both R and C++

```{r}
nm <- names(large)[length(large)]

mark(
  cppally_one_lookup = do_lookup(large, nm, 1),
  base_one_lookup = r_do_lookup(large, nm, 1)
)
```

While I'm not sure why cppally's linear scan is faster than R's, it may
be due to the fact that direct data access is used to compare elements
instead of `STRING_ELT()`

## The cost of many lookups

We will now plot the per-call cost of doing many repeated lookups to see how 
many lookups we need to do before we start seeing the benefits of hashing

```{r, fig.width=11, fig.height=7, out.width="100%", echo=FALSE}
cost_per_lookup <- numeric(10^3)
measure_ms <- function(expr){
  unclass(bench::bench_time(expr))[["real"]] * 10^3
}

for (i in 1:10^3){
  cost_per_lookup[i] <- measure_ms(do_lookup(large, nm, i)) / i
}


pt_cols <- ifelse(
    seq_along(cost_per_lookup) == 1, "orange", "black"
)
pt_cols <- ifelse(
    seq_along(cost_per_lookup) == 2, "#0072B2", pt_cols
)

plot(
    cost_per_lookup, 
    xlab = "N lookups",
    ylab = "Cost per lookup (ms)",
    col = pt_cols
)
points(1, cost_per_lookup[1], pch = 19, col = "orange")
points(2, cost_per_lookup[2], pch = 19, col = "#0072B2")
abline(h = cost_per_lookup[1], lty = 2, col = "orange")
legend(
    "topright", 
    legend = c("1st lookup (linear scan)", "2nd lookup (hash build)"),
    col = c("orange", "#0072B2"), 
    pch = 19
)
symbols(1, cost_per_lookup[1], circles = 1, add = TRUE, inches = 0.1, fg = "orange", bg = NA)
symbols(2, cost_per_lookup[2], circles = 1, add = TRUE, inches = 0.1, fg = "#0072B2", bg = NA)
```

The plot shows an interesting story - the first lookup (linear scan)
sets a baseline cost. The second lookup pays the cost of the hash map
build, making it the most expensive per-lookup. From there, each
additional lookup amortises the one-time hash build cost over more
calls, so the average drops and stabilises well below the first-lookup
baseline.

Let's see how this compares to a pseudo hash-on-1st lookup method. 
To achieve this we will set the first lookup to be a fast linear scan of 
the first name which means the rest of real lookups use hashing.

```{r}
cpp_source(code = '
#include <cppally.hpp>
using namespace cppally;

[[cppally::register]]
r_vec<r_sexp> do_first_lookup_hashed(r_vec<r_sexp> x, r_str name, int n_iterations){
  r_vec<r_sexp> out(n_iterations);
  
  // Initial lookup as fast linear scan to force all other lookups to be hashed
  static_cast<void>(x.get(x.names().get(0)));
  
  for (int i = 0; i < n_iterations; ++i){
    out.set(i, x.get(name));
  }
  return out;
}
')
```

```{r, fig.width=11, fig.height=7, out.width="100%", echo=FALSE}
cost_per_lookup1 <- numeric(250)
cost_per_lookup2 <- numeric(250)

for (i in 1:250){
  cost_per_lookup1[i] <- measure_ms(do_lookup(large, nm, i)) / i
  cost_per_lookup2[i] <- measure_ms(do_first_lookup_hashed(large, nm, i)) / i
}

plot(
    cost_per_lookup1, 
    xlab = "N lookups",
    ylab = "Cost per lookup (ms)",
    col = "darkblue",
    type = "l"
)
lines(cost_per_lookup2, col = "darkorange")
legend(
    "topright", 
    legend = c("Hash on 2nd lookup", "Hash on 1st lookup"),
    col = c("darkblue", "darkorange"),
    lty = 1
)
```

The two lines are overlapping after a few lookups, signifying that there is 
little penalty to waiting until the 2nd lookup to build the hash map.


Let's also plot some comparisons comparing repeated hash lookups against 
repeated linear scan lookups. At the same time we will also use a larger list.

```{r}
large <- as.list(sample.int(5e05))
names(large) <- paste0("name_", seq_along(large))

cpp_source(code = '
#include <cppally.hpp>
using namespace cppally;

[[cppally::register]]
r_vec<r_sexp> do_linear_lookup(r_vec<r_sexp> x, r_str name, int n_iterations){
  r_vec<r_sexp> out(n_iterations);
  r_vec<r_str> names = as<r_vec<r_str>>(x.names());
  int n = names.length();
  auto *p = names.data(); // Use ptr to allow O2 optimisations here for fair comparison
  for (int i = 0; i < n_iterations; ++i){
    for (int j = 0; j < n; ++j){
      if (unwrap(name) == p[j]){
        out.set(i, x.view(j));
        break;
      }
    }
  }
  return out;
}
')
```


```{r, fig.width=11, fig.height=7, out.width="100%", echo=FALSE}
cost_per_hash_lookup <- numeric(250)
nm <- names(large)[length(large) %/% 2]
for (i in 1:length(cost_per_hash_lookup)){
  cost_per_hash_lookup[i] <- measure_ms(do_lookup(large, nm, i)) / i
}

nm <- names(large)[length(large)]
cost_per_linear_lookup_worst_case <- numeric(250)
for (i in 1:length(cost_per_linear_lookup_worst_case)){
  cost_per_linear_lookup_worst_case[i] <- measure_ms(do_linear_lookup(large, nm, i)) / i
}

nm <- names(large)[1]
cost_per_linear_lookup_best_case <- numeric(250)
for (i in 1:length(cost_per_linear_lookup_best_case)){
  cost_per_linear_lookup_best_case[i] <- measure_ms(do_linear_lookup(large, nm, i)) / i
}

nms <- sample(names(large))
cost_per_linear_lookup_mixed_case <- numeric(250)
for (i in 1:length(cost_per_linear_lookup_mixed_case)){
  cost_per_linear_lookup_mixed_case[i] <- measure_ms(do_linear_lookup(large, nms[i], i)) / i
}


# Plot from before

plot(
    cost_per_hash_lookup,
    xlab = "N lookups",
    ylab = "Cost per lookup (ms)",
    col = "#0072B2"
)

lines(cost_per_linear_lookup_best_case, col = "purple")
lines(cost_per_linear_lookup_worst_case, col = "red")
lines(cost_per_linear_lookup_mixed_case, col = "brown")

legend(
    "topright", 
    legend = c(
        "Lazy-hash on 2nd lookup", 
        "Linear scan: best case (name always found at start)",
        "Linear scan: worst case (name always found at end)",
        "Linear scan: mixed case (name at random location)"
    ),
    col = c("#0072B2", "purple", "red", "brown"),
    pch = c(19, NA, NA, NA),
    lty = c(NA, 1, 1, 1)
)
```

The best and worst case linear scans are presented for illustrative 
purposes even though they are not realistic with real data. The most informative 
comparison is between the trend of the 
hash lookup cost versus the linear scan (mixed case). We can see that for the
first few lookups the linear scan dominates (as expected), 
but as the number of lookups increases, the hash approach begins to outperform 
the linear scan (mixed case) and stabilises well below it.

In summary, cppally's lazy hashing trades a one-time hash build cost for 
fast repeated lookups. The approach works best in C++, 
where the cache persists across calls. 
From R, each call starts fresh, so the benefit is limited to functions that 
perform multiple lookups within a single call.

## Future development

While not explored in this vignette, it is possible to preserve the cache 
across the R/C++ boundary. By attaching an `externalptr` to the `names_map` 
as an attribute of a custom R vector class, the cache lifetime is tied to the 
lifetime of the R object (provided that `externalptr` is kept as an attribute).
This may require cppally support for injecting a pre-built cache into a 
fresh `r_vector` wrapper, but is otherwise a tractable extension.
